---
title: "Feature Engineering"
output: github_document
---

# Creating new features
Calculate building areas, road lengths, and number of trees surrounding PurpleAir sensors. Create new columns to represent temporal aspects such as day, hour, and weekend.

```{r setup, include=FALSE}
preprocessing_directory <- readr::read_file("inputs/preprocessing_directory.txt")
```

## Load required libraries
```{r, load-libraries, message = FALSE, warning = FALSE}
library(dplyr) # For data manipulation
library(data.table) # Faster than dataframes (for big files)
library(sf) # For working with spatial data
# library(mapview) # For interactive maps
library(lubridate) # Dates
# library(purrr) 
```

## Read files
```{r, read-files, eval=FALSE}
purpleair_data <- fread(paste0(preprocessing_directory,"/purpleair_2018-2019.csv"))
purpleair_sensors <- st_read(paste0(purpleair_directory, "/purpleair_sensors.gpkg"), quiet = TRUE)
# osm_roads <- st_read(paste0(osm_directory, "/sanfrangrid_roads_osm.gpkg"), quiet = TRUE)
```

# filter temps to year and month and create datetime column
```{r filter temps}
temps <- temps %>% mutate(Date = ymd(Date)) %>% filter(year(Date) == year, month(Date) == month) %>%
  mutate(datetime = ymd_hms(paste(Date, Time))) %>%
  select(datetime, everything(), -Date, -Time)
```

# Get area of buildings for each PA sensor
```{}
path = '/Users/heba/Desktop/Uni/Lim Lab/uber_purpleair/'
purpleair_buildings <- st_read(paste0(path,"purpleair_buildings.shp"))

purpleair_building_areas <- purpleair_buildings %>%
  group_by(sensr_d, type) %>%
  summarize(total_area = sum(st_area(geometry)))

st_write(purpleair_building_areas, "/Users/heba/Desktop/Uni/Lim Lab/uber_purpleair/purpleair_building_areas.shp", append=FALSE)

```

# Road lengths
```{r, eval=FALSE}
path = '/Users/heba/Desktop/Uni/Lim Lab/uber_purpleair/'
purpleair_uber_roads <- st_read(paste0(path,'purpleair_uber_roads.shp'))

purpleair_road_length <- purpleair_uber_roads %>%
  group_by(sensr_d, type) %>%
  summarize(road_length = sum(st_length(geometry)))

st_write(purpleair_road_length, "/Users/heba/Desktop/Uni/Lim Lab/uber_purpleair/purpleair_road_length.shp", append=FALSE)
```


# Create date variables for prediction
```{r}
result$local_timestamp <- with_tz(result$time_stamp, tzone = "America/Los_Angeles")

# # no holidays in june so it doesnt make sense to use for this month
# result$local_date <- as.Date(result$local_timestamp)
# result$is_holiday <- result$local_date %in% as.Date(holidays)

result$day_of_week <- wday(result$local_timestamp)
result$hour <- hour(result$local_timestamp)
result$day <- day(result$local_timestamp)
result$month <- month(result$local_timestamp)
result$year <- year(result$local_timestamp)
result$is_weekend <- ifelse(result$day_of_week %in% c(6, 7), 1, 0) # 6 and 7 represent the weekend

head(result)
```
