---
title: "Feature Engineering"
output: github_document
---

# Creating new features
Calculate building areas, road lengths, and number of trees surrounding PurpleAir sensors. Create new columns to represent temporal aspects such as day, hour, and weekend.

```{r setup, include=FALSE}
preprocessing_directory <- readr::read_file("inputs/preprocessing_directory.txt")
osm_directory <- readr::read_file("inputs/osm_directory.txt")
```

## Load required libraries

```{r, load-libraries, message = FALSE, warning = FALSE}
library(dplyr) # For data manipulation
library(data.table) # Faster than dataframes (for big files)
library(sf) # For working with spatial data
library(lubridate) # Dates
library(tidyr) # Pivot data
library(leaflet) # Interactive maps
library(ggplot2) # Plots
library(timeDate) # For holidays
```

## Read files

```{r, read-files}
traffic_data <- fread(paste0(preprocessing_directory, "/traffic.csv"))
purpleair_data <- fread(paste0(preprocessing_directory,"/purpleair_filtered_2018-2019.csv"))
weather_data <- fread(paste0(preprocessing_directory,"/weather_filtered.csv"))
purpleair_sensors <- st_read(paste0(preprocessing_directory, "/pasensors_weatherstations.gpkg"), quiet = TRUE)
osm_roads <- st_read(paste0(osm_directory, "/bayarea_roads_osm.gpkg"), quiet = TRUE)
osm_trees <- st_read(paste0(osm_directory, "/bayarea_trees_osm.gpkg"), quiet = TRUE)
osm_buildings <- st_read(paste0(osm_directory, "/bayarea_buildings_osm.gpkg"), quiet = TRUE)
```

```{r, filter-for-testing, eval = FALSE, include = FALSE}
bbox <- c(xmin = -122.55, ymin = 37.82, xmax = -122.35, ymax = 37.7)
bbox_polygon <- st_as_sfc(st_bbox(bbox))
st_crs(bbox_polygon) <- 4326
sanfran <- st_intersection(osm_roads, bbox_polygon)

traffic_data <- traffic_data %>% filter(osm_way_id %in% unique(sanfran$osm_id))
purpleair_data <- purpleair_data %>% filter(sensor_index %in% unique(sanfran$sensor_index))
osm_roads <- osm_roads %>% filter(sensor_index %in% unique(sanfran$sensor_index))
osm_trees <- osm_trees %>% filter(sensor_index %in% unique(sanfran$sensor_index))
osm_buildings <- osm_buildings %>% filter(sensor_index %in% unique(sanfran$sensor_index))
purpleair_sensors <- purpleair_sensors %>% filter(sensor_index %in% unique(sanfran$sensor_index))
weather_data <- weather_data %>% filter(station %in% unique(purpleair_sensors$weatherstation))

# When testing save files to different folder
preprocessing_directory <- paste0(preprocessing_directory,"/test")
st_write(purpleair_sensors, paste0(preprocessing_directory, "/pasensors_weatherstations.gpkg"), append=FALSE, quiet = TRUE)
```

## Round sensor coordinates to 3 decimal places

```{r, round-3dp}
# Round lat and lon
purpleair_sensors <- st_set_precision(purpleair_sensors, precision=1000)
st_write(purpleair_sensors, paste0(preprocessing_directory, "/pasensors_weatherstations3dp.gpkg"), append=FALSE, quiet = TRUE)
```

## Map difference in precision

```{r, map-3dp}
purpleair_original <- st_read(paste0(preprocessing_directory, "/pasensors_weatherstations.gpkg"), quiet = TRUE)
purpleair_sensors <- st_read(paste0(preprocessing_directory, "/pasensors_weatherstations3dp.gpkg"), quiet = TRUE)

# Map difference in precision
leaflet() %>%
  addProviderTiles("CartoDB") %>% 
  addCircleMarkers(data = purpleair_original, color = "red",radius=2, label=purpleair_original$sensor_index) %>%
  addCircleMarkers(data = purpleair_sensors, color = "blue",radius=2, label=purpleair_sensors$sensor_index) %>% 
  addLegend(colors = c("red", "blue"), 
            labels = c("Original Coord", "Rounded 3dp")) %>% 
  setView(lng = -122.44, lat =  37.76, zoom = 13)
```

## Create cartesian coordinates for purple air sensors

```{r, geo-features}
# Create features to represent lat and lon
# https://medium.com/@manishsingh7163/converting-from-latitude-longitude-to-cartesian-coordinates-9ddf30841c45

# Extract coordinates
coords <- st_coordinates(purpleair_sensors)

# Convert coordinates from degrees to radians
lat_rad <- coords[, 2] * pi / 180
lon_rad <- coords[, 1] * pi / 180

# Calculate Cartesian coordinates
x <- cos(lat_rad) * cos(lon_rad)
y <- cos(lat_rad) * sin(lon_rad)
z <- sin(lat_rad)

# Add Cartesian coordinates as new columns to the purpleair_sensors data frame
purpleair_sensors <- purpleair_sensors %>%
  mutate(x = x, y = y, z = z) %>% 
  st_drop_geometry()

# Save purple air sensors
fwrite(purpleair_sensors, paste0(preprocessing_directory,"/purpleair_sensors.csv"))

# Get unique sensor indices from both data frames
unique_sensors_padata <- unique(purpleair_data$sensor_index)
unique_sensors_sensordata <- unique(purpleair_sensors$sensor_index)

# Find the intersection of the two sets of unique sensor indices
unique_sensors <- intersect(unique_sensors_padata, unique_sensors_sensordata)

# Number of purple air sensors
num_sensors <- length(unique_sensors)
```

## Weather data

```{r, join-weather}
# Drop lon and lat from weather data
weather_data <- weather_data %>% select(-lon, -lat)

# Join weather data with purpleair_sensors to get the weatherstation-related data
weather_data <- purpleair_sensors %>%
  left_join(weather_data, by = c("weatherstation" = "station"))

# Convert station to numeric as factor
weather_data$weatherstation <- as.numeric(as.factor(weather_data$weatherstation))

# Save weather data
fwrite(weather_data, paste0(preprocessing_directory,"/weather_data.csv"))
```

```{r, include=FALSE}
# Remove to free memory
rm(unique_sensors_padata,unique_sensors_sensordata,x,y,z,lat_rad,lon_rad)
```

## Road types: Major and Minor

```{r, road-types}
# roads in purple air dataset
osm_roads_pa <- osm_roads %>% filter(sensor_index %in% unique_sensors)

# https://wiki.openstreetmap.org/wiki/Key:highway
major_roads <- c("motorway", "motorway_link", "trunk", "primary")

# group road types to major and minor
osm_roads_pa <- osm_roads_pa %>%
  mutate(highway = ifelse(is.na(highway), "NA", highway)) %>%
  mutate(road_type = ifelse(highway %in% major_roads, "major", "minor"))
```

## Calculate road lengths for each PurpleAir sensor

```{r, road-lengths}
# Calculate road lengths
road_lengths <- osm_roads_pa %>%
  select(sensor_index, road_type) %>%
  group_by(sensor_index, road_type) %>%
  summarize(road_length = sum(st_length(geom)), .groups = 'drop') %>%
  st_drop_geometry()
  
# Remove units from road_length
attributes(road_lengths$road_length) = NULL

# Round road length
road_lengths$road_length <- round(road_lengths$road_length,2)

# Pivot road lengths
road_lengths_pivot <- road_lengths %>%
  pivot_wider(
    names_from = road_type,
    values_from = road_length,
    values_fill = list(road_length = 0),
    names_prefix = "length_"
  )

# Save road lengths
fwrite(road_lengths_pivot, paste0(preprocessing_directory,"/road_lengths_pivot.csv"))
```

## Traffic preprocessing

```{r, traffic-preprocessing}
# roads dataframe to join with traffic data
osm_road_types <- osm_roads_pa %>%
  select(sensor_index, osm_id, road_type) %>%
  mutate(osm_id = as.integer(osm_id)) %>%
  st_drop_geometry() %>%
  distinct()

# Save road types
fwrite(osm_road_types, paste0(preprocessing_directory,"/osm_road_types.csv"))

# Drop free flow speed since we have the speed and congestion ratio 
traffic_data <- traffic_data %>% select(-free_flow_speed)

# inner join to only keep roads that have a sensor index
traffic_roads <- traffic_data %>% inner_join(osm_road_types, by=c("osm_way_id"="osm_id"))

# Remove to free memory
rm(traffic_data,osm_road_types,osm_roads_pa,osm_roads)

# Aggregate traffic data by sensor, road_type and timestamp
traffic_agg <- traffic_roads %>%
  group_by(sensor_index, road_type, utc_timestamp) %>%
  summarize(
    mean_speed = mean(speed_mph_mean, na.rm = TRUE),
    median_speed = median(speed_mph_mean, na.rm = TRUE),
    mean_congestion = mean(congestion_ratio, na.rm = TRUE),
    median_congestion = median(congestion_ratio, na.rm = TRUE),
    .groups = 'drop'
  )

# Remove to free memory
rm(traffic_roads)

# Pivot the data for all relevant columns
traffic_agg <- traffic_agg %>% 
  pivot_wider(names_from = road_type, 
              values_from = c(mean_speed, median_speed, mean_congestion, median_congestion),
              names_sep = "_")

# Save traffic agg
fwrite(traffic_agg, paste0(preprocessing_directory,"/traffic_agg.csv"))
```

## Building types: 

```{r, road-types}
# # buildings in purple air dataset
# osm_buildings_pa <- osm_buildings %>% filter(sensor_index %in% unique_sensors)

# https://wiki.openstreetmap.org/wiki/Key:building
# Define building categories
house <- c("house", "detached", "semidetached_house", "houses", "farm")
apartments <- c("residential", "apartments")

# Categorize buildings
buildings <- osm_buildings %>% 
  filter(sensor_index %in% unique_sensors) %>%
  mutate(building = ifelse(is.na(building), "NA", building)) %>%
  mutate(building_type = case_when(
    building %in% house ~ "house",
    building %in% apartments ~ "apartments",
    TRUE ~ "other"
  ))
```

## Calculate building areas for each PurpleAir sensor

```{r, building-areas, eval=TRUE}
# building areas in m^2
building_areas <- buildings %>%
  select(sensor_index, building_type) %>%
  group_by(sensor_index, building_type) %>%
  summarize(building_area = sum(st_area(geom)), .groups = 'drop') %>%
  st_drop_geometry()

# remove units from building_area
attributes(building_areas$building_area) = NULL

# round area
building_areas$building_area <- round(building_areas$building_area,2)

# pivot building types
building_areas <- building_areas %>%
  pivot_wider(names_from = building_type, 
              values_from = building_area, 
              names_prefix = "area_",
              values_fill = list(building_area = 0))

# Save building areas
fwrite(building_areas, paste0(preprocessing_directory,"/building_areas.csv"))
```

## Calculate num trees for each PurpleAir sensor
```{r, tree-counts}
# Drop geometry and calculate tree counts
tree_counts <- osm_trees %>% 
  filter(sensor_index %in% unique_sensors) %>%
  st_drop_geometry() %>%
  select(sensor_index) %>%
  group_by(sensor_index) %>%
  summarize(num_trees = n(), .groups = 'drop')

# Save tree counts
fwrite(tree_counts, paste0(preprocessing_directory,"/tree_counts.csv"))
```

## Create final dataset temporal features

```{r, temporal-features}
# Get holidays for 2018 and 2019
holidays <- as.Date(c(holidayNYSE(2019), holidayNYSE(2018)))

# Create temporal features
final_dataset <- purpleair_data %>%
  filter(sensor_index %in% unique_sensors) %>%
  mutate(
    local_timestamp = with_tz(time_stamp, tzone = "America/Los_Angeles"),
    local_date = as.Date(local_timestamp, tz="America/Los_Angeles"),
    dow = lubridate::wday(local_timestamp),
    hour = lubridate::hour(local_timestamp),
    day = lubridate::day(local_timestamp),
    month = lubridate::month(local_timestamp),
    year = lubridate::year(local_timestamp),
    wknd = ifelse(dow %in% c(6, 7), 1, 0),
    holiday = ifelse(local_date %in% holidays, 1, 0)
  ) %>% select(-local_timestamp, -local_date)

# Remove to free memory
rm(purpleair_data)
```

## Join datasets
```{r, join-datasets}
# read files
traffic_agg <- fread(paste0(preprocessing_directory,"/traffic_agg.csv"))
osm_road_types <- fread(paste0(preprocessing_directory,"/osm_road_types.csv"))
road_lengths_pivot <- fread(paste0(preprocessing_directory,"/road_lengths_pivot.csv"))
weather_data <- fread(paste0(preprocessing_directory,"/weather_data.csv"))
purpleair_sensors <- fread(paste0(preprocessing_directory,"/purpleair_sensors.csv"))
tree_counts <- fread(paste0(preprocessing_directory,"/tree_counts.csv"))

# Merge road lengths with final_dataset
final_dataset <- final_dataset %>%
  left_join(road_lengths_pivot, by = "sensor_index") %>% 
  mutate(across(starts_with("length_"), ~ replace_na(.x, 0)))

# Merge traffic data with final dataset
final_dataset <- final_dataset %>%
  left_join(traffic_agg, by = c("time_stamp" = "utc_timestamp", "sensor_index" = "sensor_index"))

# Merge weather with final dataset
final_dataset <- final_dataset %>%
  left_join(weather_data, by = c("sensor_index" = "sensor_index", "time_stamp" = "timestamp"))

# Merge building counts with final_dataset
final_dataset <- final_dataset %>%
  left_join(building_areas, by = "sensor_index") %>%
  mutate(across(starts_with("area_"), ~ replace_na(.x, 0)))

# Merge tree counts with final_dataset
final_dataset <- final_dataset %>%
  left_join(tree_counts, by = "sensor_index") %>% 
  mutate(num_trees = replace_na(num_trees, 0))
```

## Save final dataset
```{r, final-dataset}
fwrite(final_dataset, paste0(preprocessing_directory,"/final_dataset.csv"))
```