---
title: "OpenStreetMap Data"
output: github_document
knit: (function(input, ...) {rmarkdown::render(input, output_dir = "../docs", )})
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(fig.path = "../docs/plots/")
```

Load required libraries

```{r, load-libraries, message = FALSE, warning = FALSE}
library(osmdata)       # Download OpenStreetMap Data
library(dplyr)         # Data manipulation
library(data.table)    # Working with large files
library(sf)            # Spatial data manipulation
library(leaflet)       # Interactive maps
library(tigris)        # Counties map data
library(ggplot2)       # Data visualization
library(htmlwidgets)   # Creating HTML widgets
library(webshot)       # Convert URL to image
library(DataOverviewR)
library(tidyverse)
library(terra)
library(tidyterra)
library(tibble)
library(here)          # Robust file paths
```

```{r, read-sensors-csv, echo = FALSE}
# Read sensors data
pa_sensors <- read.csv(here("code", "data", "raw", "pa_sensors.csv"))
pa_sf <- st_as_sf(pa_sensors, coords=c("longitude", "latitude"), crs = 4326)
```

Create 1 km octagon buffers around Purple Air Sensors

```{r, pa-buffers, eval=TRUE}
# Reproject to California Albers (meters) for buffering
pa_sf <- st_transform(pa_sf, 3310)

# Create 1 km buffers (octagons) around sensors
purpleairs_buffers <- st_buffer(pa_sf, dist=1000, nQuadSegs=2)

# Reproject back to lat/lon for mapping
pa_sf <- st_transform(pa_sf, 4326)
purpleairs_buffers <- st_transform(purpleairs_buffers, 4326)

# save buffers
st_write(purpleairs_buffers, 
         here("code", "data", "raw", "purpleairs_buffers.gpkg"),
         driver = "GPKG", append = FALSE, quiet = TRUE)
```

Map PurpleAir Sensors with Buffers

```{r, buffer-map}
img_path <- here("docs", "plots", "pa-buffer-map.png")
if (!file.exists(img_path)) {
  map_path <- here("docs", "maps", "pa-buffer-map.html")
  m <- leaflet() %>%
    addPolygons(data = purpleairs_buffers, 
                popup = ~paste("sensor_index:", sensor_index), 
                label = ~paste("sensor_index:", sensor_index),
                color = "#AA44AA", opacity = 0.5, weight = 1,
                fillColor = "#dec9e9", fillOpacity = 0.2) %>%
    addCircleMarkers(data = pa_sf, 
                     popup = ~paste("sensor_index:", sensor_index), 
                     label = ~paste("sensor_index:", sensor_index),
                     color = "#AA44AA", weight = 2, radius = 1) %>%
    addProviderTiles("CartoDB") %>%
    setView(lng = -122.4194, lat = 37.7749, zoom = 12)
  saveWidget(m, file = map_path)
  webshot(map_path, file = img_path)
}

knitr::include_graphics(img_path)
```

Download OSM Landuse, Natural, Leisure, Boundary

```{r, iferror-use-dif-server, include = FALSE, eval = FALSE}
# If osmdata is failing (Waiting 12s to retry)
# Try using different overpass server
# https://wiki.openstreetmap.org/wiki/Overpass_API#Public_Overpass_API_instances

old_url <- "https://overpass.kumi.systems/api/interpreter"
new_url <- "https://overpass-api.de/api/interpreter"
get_overpass_url()
set_overpass_url(new_url)
```

```{r, osm-land, warning = FALSE}
filepath <- here("code", "data", "raw", "bayarea_osm_land.gpkg")

process_polygons <- function(osm_data, osm_key, geom) {
  if (!is.null(osm_data) && nrow(osm_data) > 0) {
    osm_data <- st_make_valid(st_cast(osm_data, "POLYGON"))
    osm_data$osm_id <- ifelse("osm_id" %in% names(osm_data), osm_data$osm_id, rownames(osm_data))
    osm_data$name <- ifelse("name" %in% names(osm_data), osm_data$name, NA)
    # if osm_key column doesnt exist, add it
    if (!(osm_key %in% names(osm_data))) {
      osm_data[[osm_key]] <- NA_character_
    }
    if (osm_key == "boundary") {
      osm_data <- osm_data %>% filter(boundary %in% c("national_park", "protected_area"))
    }
    osm_data <- st_make_valid(osm_data)
    osm_data <- osm_data %>% mutate(osm_value = osm_data[[osm_key]], osm_key = osm_key) %>%
      select(osm_id, osm_value, osm_key, name) %>%
      st_intersection(geom)
    return(osm_data)
  }
  return(NULL)
}

if (!file.exists(filepath)) {
  osm_keys <- c("landuse", "natural", "leisure", "boundary")
  all_land_data <- NULL
  for (i in 1:nrow(purpleairs_buffers)) {
    sensor_index <- purpleairs_buffers$sensor_index[i]
    filename <- here("code", "data", "raw", "OSM", paste0("land_sensor", sensor_index, ".gpkg"))
    if (file.exists(filename)) next
    land_data <- NULL
    for (osm_key in osm_keys) {
      bb <- purpleairs_buffers[i, ]$geom
      # Download OSM data for each key
      osm <- opq(bbox = bb) %>%
        add_osm_feature(key = osm_key) %>%
        osmdata_sf()
      # Combine polygons and multipolygons
      osm_p <- process_polygons(osm$osm_polygons, osm_key, bb)
      osm_mp <- process_polygons(osm$osm_multipolygons, osm_key, bb)
      key_data <- rbind(osm_p, osm_mp)
      land_data <- rbind(land_data, key_data)
    }
    if (is.null(land_data) || nrow(land_data) == 0) next 
    land_data <- land_data %>% 
      mutate(sensor_index = sensor_index) %>%
      select(sensor_index, osm_id, osm_value, osm_key, name) 
    st_write(land_data, filename, driver = "GPKG", append = FALSE, quiet = TRUE)
  }
  # Save OSM land data for sensors into a single file
  file_paths <- list.files(here("code", "data", "raw", "OSM"),
                           pattern = "^land_sensor.*\\.gpkg$", full.names = TRUE)
  sf_list <- lapply(file_paths, st_read, quiet = TRUE)
  merged_sf <- do.call(rbind, sf_list) %>% distinct()
  st_write(merged_sf, filepath, driver = "GPKG", append = FALSE, quiet = TRUE)
}
```

```{r, land-dict}
osm_land <- st_read(here("code", "data", "raw", "bayarea_osm_land.gpkg"), quiet = TRUE, promote_to_multi = FALSE)

desc <- data_description(osm_land, 
                         var_desc = c(
                           "sensor_index" = "Unique identifier for PurpleAir sensors",
                           "osm_id" = "Unique OpenStreetMap identifier for land use feature",
                           "osm_value" = "Specific land use value/category (park, forest, industrial etc.)",
                           "osm_key" = "Top-level OSM classification (landuse, natural, leisure etc.)", 
                           "name" = "Name of the land use feature if available",
                           "geom" = "Land use polygon geometry"
                         ))

data_dictionary(osm_land, 
                data_title = "OSM Land", 
                descriptions = desc, 
                hide = c("top_n", "NA_Percentage", "NA_Count", "n_unique"))

# Show example records
knitr::kable(head(osm_land, 3),
             row.names = FALSE,
             format = "markdown",
             caption = "Example Records")
```

```{r, land-nas}
data_dictionary(osm_land, 
                data_title = "Missing Values Summary")
```

```{r, define-land-types}
osm_lands <- osm_land %>%
  filter(!is.na(osm_value)) %>%
  mutate(landtype = case_when(
    # Industrial
    osm_value %in% c("industrial", "construction", "railway", "landfill",
                     "harbour", "port", "lock", "marina", "brownfield",
                     "quarry", "military", "special_use") ~ "industrial",
    # Urban
    osm_value %in% c("commercial", "retail", "garages", "common", "civic",
                     "swimming_pool", "fitness_centre", "bowling_alley",
                     "miniature_golf", "education", "religious", "institutional") ~ "urban",
    # Residential
    osm_value %in% c("residential", "outdoor_seating") ~ "residential",
    # Natural Green
    osm_value %in% c("forest", "wood", "scrub", "heath", "grassland", "meadow",
                     "greenfield", "nature_reserve", "national_park", "conservation",
                     "protected_area", "cliff", "shrub", "shrubbery") ~ "natural_green",
    # Artificial Green
    osm_value %in% c("park", "recreation_ground", "pitch", "garden",
                     "sports_centre", "sports_hall", "stadium", "playground",
                     "golf_course", "village_green", "allotments", "cemetery",
                     "zoo", "track", "raceway", "plant_nursery", "horse_riding",
                     "fishing") ~ "artificial_green",
    # Water
    osm_value %in% c("water", "riverbank", "reservoir", "basin", "dock", "canal",
                     "pond", "salt_pond", "tidal", "marsh", "wetland", "reef",
                     "bay", "coastline") ~ "water",
    # Sparse
    osm_value %in% c("sand", "scree", "beach", "mud", "glacier", "rock",
                     "bare_rock", "farmland", "farm", "farmyard", "peninsula",
                     "vineyard") ~ "sparse",
    # Default
    TRUE ~ "other"
  ))
```

```{r, landtype-counts, echo=FALSE, eval=FALSE}
osmvcount <- osm_lands %>% st_drop_geometry() %>% 
  count(landtype, osm_key, osm_value) %>% 
  mutate(pct = paste0(round(100*(n/nrow(osm_lands)),0),"%")) %>% 
  arrange(desc(n)) %>% 
  rename(count = n)
osmvcount
write.csv(osmvcount, here("code", "data", "osmvcount.csv"), row.names = FALSE)
```


```{r, landtype-counts, echo=FALSE, eval=FALSE}
osmlcount <- osm_lands %>% 
  st_drop_geometry() %>% 
  count(landtype) %>% 
  mutate(pct = paste0(round(100*(n/nrow(osm_lands)),0),"%")) %>% 
  arrange(desc(n)) %>%
  rename(count = n)
write.csv(osmlcount, here("code", "data", "osmlcount.csv"), row.names = FALSE)
```

```{r, land-prio}
priority_order <- c("industrial", "urban", "residential", "artificial_green", 
                    "natural_green", "water", "sparse")
```

Resolve overlapping geometries

```{r, test-sensors, include=FALSE}
# # getting sensors to test on
# osm_lands %>% st_drop_geometry() %>% select(sensor_index, landtype) %>% distinct()
# list(osm_lands %>% st_drop_geometry() %>% pull(sensor_index) %>% unique())
```

```{r, land-colors}
# Define a color palette for landtype
landuse_colors <- colorFactor(
  palette = c("#90EE90", "#FF0000", "#228B22", "#800080", 
              "#D3D3D3", "#D2B48C", "#FFA500", "#0000FF"),
  domain = c("artificial_green", "industrial", "natural_green", "other", 
             "residential", "sparse", "urban", "water"))
```

```{r, land-type-map}
osm_landuse1 <- osm_lands %>% filter(sensor_index %in% c(38683, 37693, 19471, 37311, 29865, 19299))
# osm_landuse1 <- osm_lands %>% filter(sensor_index %in% c(34637, 39227, 10092, 37347, 19571, 36109))

# Map of original land
leaflet(osm_landuse1) %>%
  addProviderTiles("CartoDB") %>% 
  addPolygons(
    fillColor = ~landuse_colors(landtype), fillOpacity = 0.3, opacity = 0,
    popup = ~paste("Landuse:", landtype, ", sensor: ", sensor_index),
    label = ~paste("Landuse:", landtype, ", sensor: ", sensor_index)) %>%
  addLegend(pal = landuse_colors, values = osm_landuse1$landtype)
```

```{r, dissolve-land}
land_dissolved <- osm_lands %>%
  st_make_valid() %>%
  group_by(sensor_index, landtype) %>%
  summarize(geometry = st_union(geom), .groups = 'drop')
```

```{r, dissolve-land-map}
land_dissolved1 <- land_dissolved %>% filter(sensor_index %in% c(34637, 39227, 10092, 37347, 19571, 36109))
# land_dissolved1 <- land_dissolved %>% filter(sensor_index %in% c(38683, 37693, 19471, 37311, 29865, 19299))

# Map with same type dissolved
leaflet(land_dissolved) %>%
  addProviderTiles("CartoDB") %>%
  addPolygons(
    fillColor = ~landuse_colors(landtype), fillOpacity = 0.5, opacity = 0,
    popup = ~paste("Landuse:", landtype)) %>%
  addLegend(pal = landuse_colors, values = land_dissolved1$landtype)
```

```{r, pivot-land-new, include = FALSE}
# pivot land type wider 
land_pivot <- land_dissolved %>%
  pivot_wider(names_from = landtype, values_from = landtype, values_fill = NA_character_)
```

```{r, pivot-land-new, include = FALSE}
cols <- c(industrial = NA_character_, urban = NA_character_,
          artificial_green = NA_character_, natural_green = NA_character_,
          other = NA_character_, residential = NA_character_,
          sparse = NA_character_, water = NA_character_)
land_pivot <- add_column(land_pivot, !!!cols[setdiff(names(cols), names(land_pivot))])
```

```{r}
# Function to convert and clean to only polygons
clean_to_polygons <- function(data, type_col) {
  if(nrow(data %>% filter(!!sym(type_col) == type_col)) > 0) {
    v <- vect(data %>%
                filter(!!sym(type_col) == type_col) %>%
                select(sensor_index, !!sym(type_col)))
    # Keep polygons and multipolygons
    v <- v[geomtype(v) %in% c("polygons", "multipolygons")]
    return(v)
  }
  return(NULL)
}

# Convert keeping only polygons
v_industrial <- clean_to_polygons(land_pivot, "industrial")
v_urban <- clean_to_polygons(land_pivot, "urban")
v_residential <- clean_to_polygons(land_pivot, "residential") 
v_artificial_green <- clean_to_polygons(land_pivot, "artificial_green")
v_natural_green <- clean_to_polygons(land_pivot, "natural_green")
v_water <- clean_to_polygons(land_pivot, "water")
v_sparse <- clean_to_polygons(land_pivot, "sparse")
v_other <- clean_to_polygons(land_pivot, "other")
# v_missing <- clean_to_polygons(land_pivot, "missing")
```

```{r}
process_sensor <- function(id) {
  tryCatch({
    # Initialize list of vector subsets for this sensor
    geom_list <- list(
      industrial = if(!is.null(v_industrial)) v_industrial[v_industrial$sensor_index == id],
      urban = if(!is.null(v_urban)) v_urban[v_urban$sensor_index == id],
      residential = if(!is.null(v_residential)) v_residential[v_residential$sensor_index == id],
      artificial_green = if(!is.null(v_artificial_green)) v_artificial_green[v_artificial_green$sensor_index == id],
      natural_green = if(!is.null(v_natural_green)) v_natural_green[v_natural_green$sensor_index == id],
      water = if(!is.null(v_water)) v_water[v_water$sensor_index == id],
      sparse = if(!is.null(v_sparse)) v_sparse[v_sparse$sensor_index == id],
      other = if(!is.null(v_other)) v_other[v_other$sensor_index == id],
      # missing = if(!is.null(v_missing)) v_missing[v_missing$sensor_index == id]
    )
    
    # Create empty result
    result <- NULL
    
    # Process in priority order
    for(land_type in priority_order) {
      curr_geom <- geom_list[[land_type]]
      
      if(!is.null(curr_geom) && nrow(curr_geom) > 0) {
        # For first geometry
        if(is.null(result)) {
          result <- curr_geom
          result$landtype <- land_type
        } else {
          # Erase existing areas from new geometry before adding
          new_geom <- terra::erase(curr_geom, result)
          if(!is.null(new_geom) && nrow(new_geom) > 0) {
            new_geom$landtype <- land_type
            result <- rbind(result, new_geom)
          }
        }
      }
    }
    
    # Add sensor_index if not present
    if(!is.null(result)) {
      result$sensor_index <- id
    }
    
    return(result)
    
  }, error = function(e) {
    message(sprintf("Error processing sensor %s: %s", id, e$message))
    return(NULL)
  })
}
```

```{r}
# Process all sensors
sensor_ids <- unique(land_pivot$sensor_index)
results <- lapply(sensor_ids, process_sensor)
```

```{r}
# Combine results and clean up
landuse_resolved <- do.call(rbind, results[!sapply(results, is.null)])

# Clean up and standardize output
landuse_resolved <- landuse_resolved %>%
  select(sensor_index, landtype) %>%
  st_as_sf()
```

```{r}
# landuse_resolved1 <- landuse_resolved %>% filter(sensor_index %in% c(34637, 39227, 10092, 37347, 19571, 36109))
landuse_resolved1 <- landuse_resolved

# Map with same type dissolved
leaflet(landuse_resolved1) %>%
  addProviderTiles("CartoDB") %>%
  addPolygons(
    fillColor = ~landuse_colors(landtype), fillOpacity = 0.7, opacity = 0,
    popup = ~paste("Landuse:", landtype, " Sensor: ", sensor_index)) %>%
  addLegend(pal = landuse_colors, values = landuse_resolved1$landtype)
```


```{r}
leaflet(landuse_resolved1 %>% filter(sensor_index==38683)) %>%
  addProviderTiles("CartoDB") %>%
  addPolygons(
    fillColor = ~landuse_colors(landtype), fillOpacity = 0.7, opacity = 0,
    popup = ~paste("Landuse:", landtype, " Sensor: ", sensor_index)) %>%
  addLegend(pal = landuse_colors, values = landuse_resolved1$landtype)
```


```{r}
leaflet(landuse_resolved1 %>% filter(sensor_index==37693)) %>%
  addProviderTiles("CartoDB") %>%
  addPolygons(
    fillColor = ~landuse_colors(landtype), fillOpacity = 0.7, opacity = 0,
    popup = ~paste("Landuse:", landtype, " Sensor: ", sensor_index)) %>%
  addLegend(pal = landuse_colors, values = landuse_resolved1$landtype)
```

```{r}
# Map with land resolved
m <- leaflet(landuse_resolved) %>%
  addProviderTiles("CartoDB") %>%
  addPolygons(
    fillColor = ~landuse_colors(landtype), fillOpacity = 0.7, opacity = 0,
    popup = ~paste("Landuse:", landtype)) %>%
  addLegend(pal = landuse_colors, values = landuse_resolved$landtype)

saveWidget(m, file = here("docs", "maps", "osm_resolved.html"), selfcontained = TRUE)
```

```{r}
m
```

## ISSUE WITH OVERLAPPING SENSORS HAVING DIFFERENT LAND TYPES ##

########
```{}
land_dissolved %>% st_drop_geometry() %>% select(sensor_index, landtype) %>% 
distinct() %>% count(sensor_index) %>% arrange(desc(n))
```

```{r, calc-land-areas}
# Calculate the theoretical area of your 1km octagon buffer
buffer_area <- purpleairs_buffers %>%
  filter(sensor_index == sensor_ids[1]) %>%  # take first sensor's buffer
  st_area() %>%
  as.numeric()

sf_use_s2(FALSE)  # Temporarily disable s2

# Clean and validate geometries before area calculation
landuse_areas <- landuse_resolved %>%
  st_make_valid() %>%  # Fix invalid geometries
  st_buffer(0) %>%     # Clean up any remaining issues
  st_make_valid() %>%  # Make sure everything is valid after buffering
  mutate(
    area = as.numeric(units::set_units(st_area(geometry), "m^2"))
  )
sf_use_s2(TRUE)   # Re-enable s2

# Area calculations per land type
land_data <- landuse_areas %>%
  st_drop_geometry() %>%
  group_by(sensor_index) %>%
  summarize(
    industrial = round(sum(area[landtype == "industrial"], na.rm = TRUE)),
    urban = round(sum(area[landtype == "urban"], na.rm = TRUE)),
    residential = round(sum(area[landtype == "residential"], na.rm = TRUE)),
    artificial_green = round(sum(area[landtype == "artificial_green"], na.rm = TRUE)),
    natural_green = round(sum(area[landtype == "natural_green"], na.rm = TRUE)),
    water = round(sum(area[landtype == "water"], na.rm = TRUE)),
    sparse = round(sum(area[landtype == "sparse"], na.rm = TRUE)),
    other = round(sum(area[landtype == "other"], na.rm = TRUE)),
    # missing = round(sum(area[landtype == "missing"], na.rm = TRUE))
  )

# Optional: Calculate percentage of total area for each land type
land_data_pct <- land_data %>%
  rowwise() %>%
  mutate(
    total_area = buffer_area,
    across(-c(sensor_index, total_area), 
           ~round(. / total_area, 2))) %>%
  select(-total_area)

write.csv(land_data_pct, file = here("code", "data", "processed", "land_data.csv"), row.names = FALSE)
```


```{}
# # Temporarily disable S2 geography
# sf_use_s2(FALSE)
# 
# clean_result <- landuse_resolved %>%
#   st_make_valid() %>%
#   st_buffer(0) %>%
#   st_make_valid() %>%
#   st_cast("MULTIPOLYGON")

# Calculate areas
landuse_areas <- clean_result %>%
mutate(
area = as.numeric(st_area(geometry))
)
# 
# # Re-enable S2 if needed for other operations
# sf_use_s2(TRUE)

# area calc
land_data <- landuse_areas %>%
st_drop_geometry() %>%
group_by(sensor_index) %>%
summarize(
industrial = round(sum(area[landtype == "industrial"], na.rm = TRUE)),
urban = round(sum(area[landtype == "urban"], na.rm = TRUE)),
artificial = round(sum(area[landtype == "artificial"], na.rm = TRUE)),
natural = round(sum(area[landtype == "natural"], na.rm = TRUE)),
missing = round(sum(area[landtype == "missing"], na.rm = TRUE))
)
```

```{r, save-land-data}
# write.csv(land_data, file = here("code", "data", "processed", "land_data.csv"), row.names = FALSE)
```







Frequency of osm keys and values near sensors

```{r}
key_value_counts <- osm_land %>%
  st_drop_geometry() %>%
  distinct() %>%
  group_by(osm_key, osm_value) %>%
  summarize(n_osmids = n_distinct(osm_id),
            n_sensors = n_distinct(sensor_index),
            .groups = 'drop') %>% 
  arrange(osm_key, desc(n_sensors))
key_value_counts
```

```{r}
key_counts <- osm_land %>%
  st_drop_geometry() %>%
  distinct() %>%
  group_by(osm_key) %>%
  summarize(n_osmids = n_distinct(osm_id),
            n_sensors = n_distinct(sensor_index),
            .groups = 'drop') %>% 
  arrange(osm_key, desc(n_sensors))
key_counts
```

Check where NA values

```{r}
library(mapview)
osm_land %>% filter(is.na(osm_value))


na_check <- osm_land %>% filter(osm_id == 548274970, sensor_index == 10092)
mapview(na_check, col.regions = "#8ac926", alpha.regions = 0.5)
```

Might need later

```{r, eval = FALSE, echo = FALSE}
(is.na(access) | access != "private")
(key == "leisure", land == "swimming_pool")

pools <- osm_land %>% filter(osm_key == "leisure", osm_value == "swimming_pool")
```


Map of polygon outlines and polygons filled
With labels (hover) and popups (click) 

```{r, eval = FALSE, echo = FALSE}
colors = c("#1982c4", "#8ac926", "#ff595e")

m <- leaflet() %>%
  addPolygons(data = purpleairs_buffers[i, ],
              popup = ~paste("sensor_index:", sensor_index),
              label = ~paste("sensor_index:", sensor_index),
              color = "black", opacity = 1, weight = 1,
              fillOpacity = 0) %>%
  addPolygons(data = osm_data,
              # popup = ~paste(name, "osm_id:", osm_id),
              # label = ~paste(name, "osm_id:", osm_id),
              color = "#8ac926", opacity = 0.5, weight = 2, 
              fillColor = "#8ac926", fillOpacity = 0) %>%
  addProviderTiles("CartoDB")
m
```

Mapview

```{r}
mapview(onlynatural, col.regions = "#8ac926", alpha.regions = 0.5) + mapview(onlyboundary, col.regions = "#1982c4", alpha.regions = 0.5)
```

Save map

```{r, eval = FALSE, echo = FALSE}
img_path <- here("docs", "plots", "landuse-map.png")
if (!file.exists(img_path)) {
  map_path <- here("docs", "maps", "landuse-map.html")
  m <- leaflet() 
  # save map as html
  saveWidget(m, file = map_path)
  # save html as image
  webshot(map_path, file = img_path)
}
knitr::include_graphics(img_path)
```

Select columns with at least 90% data 

```{r, eval = FALSE, echo = FALSE}
x <- x %>% select(where(~ sum(is.na(.)) / nrow(x) < 0.9))
```






Map of OpenStreetMap Buildings

```{r, eval = FALSE, echo = FALSE}
color <- "#D7D0CA"

bayarea_buildings <- st_read(here("code", "data", "raw", "bayarea_buildings_osm.gpkg"), quiet = TRUE)
ca <- counties("California", cb = TRUE, progress_bar=FALSE)

ggplot() + 
  geom_sf(data = ca, color="black", fill="antiquewhite", size=0.25) +
  geom_sf(data = bayarea_buildings, color="darkorange3", fill = "darkorange") +
  coord_sf(xlim = c(-123.8, -121.0), ylim = c(36.9, 39.0)) +
  theme(panel.background = element_rect(fill = "aliceblue")) + 
  xlab("Longitude") + ylab("Latitude") +
  ggtitle("OSM Buildings")
```

```{r, eval = FALSE, echo = FALSE}
## Select small area of San Francisco to map

bbox <- c(xmin = -122.46, ymin = 37.76, xmax = -122.42, ymax = 37.72)
bbox_polygon <- st_as_sfc(st_bbox(bbox))
st_crs(bbox_polygon) <- 4326
sanfran_roads <- st_intersection(bayarea_roads, bbox_polygon)
sanfran_buildings <- st_intersection(bayarea_buildings, bbox_polygon)
sanfran_trees <- st_intersection(bayarea_trees, bbox_polygon)
sanfran_sensors <- st_intersection(filtered_sensors, bbox_polygon)
sanfran_buffers <- st_intersection(purpleairs_buffers, bbox_polygon)


## Plot san fran city
leaflet() %>%
  # addPolylines(data = bbox_polygon, color = "black", weight = 3, opacity = 1, popup = "Bounding Box") %>%
  addPolylines(data = sanfran_roads, color = "cornflowerblue", weight = 1, popup = "Roads") %>%
  addPolygons(data = sanfran_buildings, fillOpacity = 1, weight = 1, color="darkorange", fillColor = "orange", popup = "Buildings") %>%
  addCircleMarkers(data = sanfran_trees, popup = "Trees",
                   fillColor = "#99CC99", fillOpacity = 1,
                   color = "#336600", weight=1, opacity = 1, radius = 1.5) %>%
  addCircleMarkers(data = sanfran_sensors, popup = "Sensors",
                   fillColor = "#CC66CC", fillOpacity = 1,
                   color = "#9933CC",weight=2, opacity = 1, radius = 5) %>%
  addPolygons(data = sanfran_buffers, color = "#9933CC", weight = 1, opacity = 1, fillOpacity = 0, popup = "Buffers") %>%
  addProviderTiles("CartoDB") %>% 
  addLegend(colors = c("#9933CC", "cornflowerblue", "darkorange", "#336600"), 
            labels = c("PurpleAir Sensors", "Roads", "Buildings", "Trees"))


## Plot san fran city


library(htmlwidgets)
library(htmltools)


major_roads <- c("motorway", "motorway_link", "trunk", "primary")

# group road types to major and minor
osm_roads_pa <- bayarea_roads %>%
  mutate(highway = ifelse(is.na(highway), "NA", highway)) %>%
  mutate(road_type = ifelse(highway %in% major_roads, "Major", "Minor"),
         road_length = round(st_length(geom),2)) 

# Define building categories
house <- c("house", "detached", "semidetached_house", "houses", "farm")
apartments <- c("residential", "apartments")
undefined <- c("NA", "yes")

# Categorize buildings
buildings <- bayarea_buildings %>% 
  mutate(building = ifelse(is.na(building), "NA", building)) %>%
  mutate(building_type = case_when(
    building %in% undefined ~ "Undefined",
    building %in% house ~ "House",
    building %in% apartments ~ "Apartments",
    TRUE ~ "Other"
  ),
  building_area = round(st_area(geom),2))

# Remove units from road_length
# attributes(road_lengths$road_length) = NULL

# Define colors for road types
road_colors <- colorFactor(c("cornflowerblue", "lightblue"), domain = c("Major", "Minor"))
building_colors <- colorFactor(c("orangered", "darkorange", "peachpuff", "papayawhip"), 
                               domain = c("Other", "Apartments", "House", "Undefined"))

m <- leaflet() %>%
  addPolylines(data = osm_roads_pa, color = ~road_colors(road_type), weight = 2,
               label = ~HTML(paste(htmlEscape(road_type), " (", htmlEscape(highway),
                                   ")<br>Length: ", htmlEscape(road_length), " [m]")),
               popup = "Roads") %>%
  addPolygons(data = buildings, fillOpacity = 1, weight = 1,
              color=~building_colors(building_type), fillColor = ~building_colors(building_type),
              label = ~HTML(paste(htmlEscape(building_type), " (", htmlEscape(building),
                                  ")<br>Area: ", htmlEscape(building_area), " [m^2]"),
                            popup = "Buildings")) %>%
  addCircleMarkers(data = bayarea_trees, popup = "Trees", label = "Tree",
                   fillColor = "#99CC99", fillOpacity = 1,
                   color = "#336600", weight=1, opacity = 1, radius = 2) %>%
  addCircleMarkers(data = z, popup = "Sensors",
                   label = ~paste("Sensor:", htmlEscape(sensor_index)),
                   fillColor = "#CC66CC", fillOpacity = 1, color = "#9933CC",
                   weight=2, opacity = 1, radius = 5) %>%
  addPolylines(data = purpleairs_buffers, color = "#9933CC", weight = 1,
               opacity = 1, popup = "Buffers") %>%
  addProviderTiles("CartoDB") %>% 
  addLegend(colors = c("#9933CC", "#336600", "cornflowerblue", "lightblue",
                       "orangered", "darkorange", "peachpuff", "oldlace"), 
            labels = c("PurpleAir Sensors", "Trees", "Major Roads", "Minor Roads",
                       "Other", "Apartments", "House", "Undefined"))
m
saveWidget(m, file="san_fran_map.html")
```


########### DEBUG ########### 
########### DEBUG ########### 
########### DEBUG ########### 

```{r}
# 38683, 37693
osm_land %>% filter(sensor_index==37693, osm_value=="industrial")
```

```{r}
osm_land %>% filter(sensor_index==38683, osm_value=="industrial")
```

```{r}
land_data <- NULL
osm_keys <- c("landuse", "natural", "leisure", "boundary")
sensor_index <- 38683
buffer <- purpleairs_buffers %>% filter(sensor_index==38683)
filename <- here("code", "data", "raw", "OSM", paste0("land_sensor", sensor_index, ".gpkg"))
osm_key = "landuse"
bb <- buffer$geom
# Download OSM data for each key
osm <- opq(bbox = bb) %>%
  add_osm_feature(key = osm_key) %>%
  osmdata_sf()
# Combine polygons and multipolygons
osm_p <- process_polygons(osm$osm_polygons, osm_key, bb)
osm_mp <- process_polygons(osm$osm_multipolygons, osm_key, bb)
key_data <- rbind(osm_p, osm_mp)
land_data <- rbind(land_data, key_data)

if (is.null(land_data) || nrow(land_data) == 0) next 
land_data <- land_data %>% 
  mutate(sensor_index = sensor_index) %>%
  select(sensor_index, osm_id, osm_value, osm_key, name) 
```
```{r}
osm$osm_multipolygons
```

```{r}
geom<-bb
osm_data<-osm$osm_multipolygons
osm_data <- st_make_valid(st_cast(osm_data, "POLYGON"))
osm_data$osm_id <- ifelse("osm_id" %in% names(osm_data), osm_data$osm_id, rownames(osm_data))
osm_data$name <- ifelse("name" %in% names(osm_data), osm_data$name, NA)
osm_key = "landuse"
# osm_key = "testcol"
# osm_data[[osm_key]] <- ifelse(osm_key %in% names(osm_data), osm_data[[osm_key]], NA)
if (!(osm_key %in% names(osm_data))) {
  osm_data[[osm_key]] <- NA_character_  # Ensures new column is character type
}

if (osm_key == "boundary") {
  osm_data <- osm_data %>% filter(boundary %in% c("national_park", "protected_area"))
}
osm_data <- st_make_valid(osm_data)
osm_data <- osm_data %>% mutate(osm_value = osm_data[[osm_key]], osm_key = osm_key) %>%
  select(osm_id, osm_value, osm_key, name)

osm_data <- osm_data %>% st_intersection(geom)
```

```{r}
osm_land <- osm_land %>% filter(sensor_index %in% c(38683, 37693))

# Create random colors for unique `osm_value`
osm_land <- osm_land %>% mutate(random_color = sample(colors(), n(), replace = TRUE))
osm_land <- osm_land %>% mutate(random_color = ifelse(osm_value=="industrial","red","black"))
```

```{r}
# osm_data <- osm_data %>% mutate(random_color = ifelse(landuse=="construction","red","black"))

leaflet(osm_land) %>%
  addProviderTiles("CartoDB") %>% 
  addPolygons(
    fillColor = ~random_color, fillOpacity = 0.5, opacity = 0,
    popup = ~paste("osm_id: ", osm_id, "osm_value:", osm_value, ", sensor: ", sensor_index, name),
    label = ~paste("osm_id: ", osm_id, "osm_value:", osm_value, ", sensor: ", sensor_index, name))

```

```{r}
sensor38683 <- st_read(here("code", "data", "raw", "OSM", "land_sensor38683.gpkg"), quiet = TRUE)
sensor37693 <- st_read(here("code", "data", "raw", "OSM", "land_sensor37693.gpkg"), quiet = TRUE, promote_to_multi = FALSE)
unique(sensor38683$osm_value)
unique(sensor37693$osm_value)

osm_land %>% filter(sensor_index==38683)
sensor38683
```

```{r}
leaflet(osm_land %>% filter(sensor_index==38683)) %>%
  addProviderTiles("CartoDB") %>% 
  addPolygons(
    fillColor = ~random_color, fillOpacity = 0.5, opacity =0,
    popup = ~paste("osm_id: ", osm_id, "osm_value:", osm_value, ", sensor: ", sensor_index, name),
    label = ~paste("osm_id: ", osm_id, "osm_value:", osm_value, ", sensor: ", sensor_index, name))
```

```{r}
leaflet(osm_land %>% filter(sensor_index==37693)) %>%
  addProviderTiles("CartoDB") %>% 
  addPolygons(
    fillColor = ~random_color, fillOpacity = 0.5, opacity =0,
    popup = ~paste("osm_id: ", osm_id, "osm_value:", osm_value, ", sensor: ", sensor_index, name),
    label = ~paste("osm_id: ", osm_id, "osm_value:", osm_value, ", sensor: ", sensor_index, name))
```

