---
title: "Download OSM data"
output: github_document
knit: (function(input, ...) {rmarkdown::render(input, output_dir = "../docs", )})
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(fig.path = "../docs/plots/")
```

```{r, github-package, include=FALSE}
if (!"DataOverviewR" %in% rownames(installed.packages())) {
  suppressMessages({devtools::install_github("heba-razzak/DataOverviewR")})
}
```

Load required libraries

```{r, load-libraries, message = FALSE, warning = FALSE}
library(osmdata)       # Download OpenStreetMap Data
library(dplyr)         # Data manipulation
library(data.table)    # Working with large files
library(sf)            # Spatial data manipulation
library(leaflet)       # Interactive maps
library(tigris)        # Counties map data
library(ggplot2)       # Data visualization
library(htmlwidgets)   # Creating HTML widgets
library(webshot)       # Convert URL to image
```

```{r, read-sensors-csv, echo = FALSE}
# Read sensors data
filepath <- file.path("data", "raw", "pa_sensors.csv")
pa_sensors <- read.csv(filepath)
pa_sf <- st_as_sf(pa_sensors, coords=c("longitude", "latitude"), crs = 4326)
```

Create 1 km octagon buffers around Purple Air Sensors

```{r, pa-buffers, eval=TRUE}
# Reproject to California Albers (meters) for buffering
pa_sf <- st_transform(pa_sf, 3310)

# Create 1 km buffers (octagons) around sensors
purpleairs_buffers <- st_buffer(pa_sf, dist=1000, nQuadSegs=2)

# Reproject back to lat/lon for mapping
pa_sf <- st_transform(pa_sf, 4326)
purpleairs_buffers <- st_transform(purpleairs_buffers, 4326)
```

Map PurpleAir Sensors with Buffers

```{r, stations-map}
img_path <- file.path("../docs", "plots", "pa-buffer-map.png")
if (!file.exists(img_path)) {
  map_path <- file.path("../docs", "maps", "pa-buffer-map.html")
  m <- leaflet() %>%
    addCircleMarkers(data = pa_sf, popup = ~as.character(sensor_index), label = ~as.character(sensor_index),
                     color = "#AA44AA", weight = 2, radius = 2) %>%
    addPolygons(data = purpleairs_buffers, popup = ~id,
                color = "#AA44AA", opacity = 0.5, weight = 1,
                fillColor = "#dec9e9", fillOpacity = 0.2) %>%
    addProviderTiles("CartoDB") %>%
    setView(lng = -122.4194, lat = 37.7749, zoom = 12)
  saveWidget(m, file = map_path)
  webshot(map_path, file = img_path)
}

knitr::include_graphics(img_path)
```

Download OSM Roads

```{r, download-roads, warning=FALSE}
filepath <- file.path("data", "raw", "bayarea_osm_roads.gpkg")

if (!file.exists(filepath)) {
  # Download OSM roads data for each sensor buffer
  for (i in 1:nrow(purpleairs_buffers)) {
    output_name <- paste0("sensor", purpleairs_buffers$sensor_index[i], "_osm_roads.gpkg")
    filename <- file.path("data", "raw", "OSM", output_name)
    
    if (file.exists(filename)) next
    
    osm <- opq(bbox = purpleairs_buffers[i, ]$geom) %>%
      add_osm_feature(key = 'highway') %>%
      osmdata_sf()
    
    if (is.null(osm$osm_lines) || nrow(osm$osm_lines) == 0) next
    
    # If OSM id is missing, fill with row names
    if(!"osm_id" %in% names(osm$osm_lines)) {
      osm$osm_lines$osm_id <- rownames(osm$osm_lines)
    }
    
    # If other column is missing, fill with NA
    cols <- c("name", "highway", "lanes", "maxspeed")
    for (c in cols) {
      if(!c %in% names(osm$osm_lines)) {
        osm$osm_lines[[c]] <- NA
      }
    }
    
    selected_columns <- osm$osm_lines %>%
      select(osm_id, name, highway, lanes, maxspeed)
    
    # Intersect with buffer
    sf_obj <- st_intersection(st_as_sf(selected_columns), purpleairs_buffers[i, ]$geom)
    
    if (is.null(sf_obj) || nrow(sf_obj) == 0) next
    
    sf_obj$sensor_index <- purpleairs_buffers$sensor_index[i]
    st_write(sf_obj, filename, driver = "GPKG", append = FALSE, quiet = TRUE)
  }
  
  # Save OSM road data for sensors into a single file
  file_paths <- list.files(file.path("data", "raw", "OSM"), pattern = "^sensor.*_osm_roads\\.gpkg$", full.names = TRUE)
  sf_list <- lapply(file_paths, st_read, quiet = TRUE)
  merged_sf <- do.call(rbind, sf_list) %>% distinct()
  st_write(merged_sf, filepath, driver = "GPKG", append = FALSE, quiet = TRUE)
}
```

Download OSM Landuse, Natural, Leisure, Boundary, Waterway

```{r, osm-land, warning = FALSE}
filepath <- file.path("data", "raw", "bayarea_osm_land.gpkg")

process_polygons <- function(osm_data, land_key, geom) {
  if (!is.null(osm_data) && nrow(osm_data) > 0) {
    osm_data <- st_make_valid(st_cast(osm_data, "POLYGON"))
    osm_data$osm_id <- ifelse("osm_id" %in% names(osm_data), osm_data$osm_id, rownames(osm_data))
    osm_data[[land_key]] <- ifelse(land_key %in% names(osm_data), osm_data[[land_key]], NA)
    if (land_key == "boundary") {
      osm_data <- osm_data %>% filter(boundary %in% c("national_park", "protected_area"))
    }
    osm_data <- osm_data %>% mutate(land = osm_data[[land_key]], land_key = land_key) %>%
      select(osm_id, land, land_key) %>%
      st_intersection(geom)
    return(osm_data)
  }
  return(NULL)
}

if (!file.exists(filepath)) {
  osm_keys <- c("landuse", "natural", "leisure", "boundary")
  all_land_data <- NULL
  # for (i in 1:nrow(purpleairs_buffers)) {
  for (i in 1:1) {
    sensor_index <- purpleairs_buffers$sensor_index[i]
    land_data <- NULL
    for (land_key in osm_keys) {
      bb <- purpleairs_buffers[i, ]$geom
      # Download OSM data for each key
      osm <- opq(bbox = bb) %>%
        add_osm_feature(key = land_key) %>%
        osmdata_sf()
      # Combine polygons and multipolygons
      osm_p <- process_polygons(osm$osm_polygons, land_key, bb)
      osm_mp <- process_polygons(osm$osm_multipolygons, land_key, bb)
      key_data <- rbind(osm_p, osm_mp)
      land_data <- rbind(land_data, key_data)
    }
    if (is.null(land_data) || nrow(land_data) == 0) next 
    land_data <- land_data %>% 
      mutate(sensor_index = sensor_index) %>%
      select(sensor_index, osm_id, land, land_key) 
    all_land_data <- rbind(all_land_data, land_data)
  }
  st_write(all_land_data, filepath, driver = "GPKG", append = FALSE, quiet = TRUE)
}
```

```{r}
mapview(all_land_data)
```

```{r}
filepath <- file.path("data", "raw", "bayarea_osm_roads.gpkg")
osm_roads <- st_read(filepath)
# filepath <- file.path("data", "raw", "bayarea_osm_land_no_intersection.gpkg")
filepath <- file.path("data", "raw", "bayarea_osm_land.gpkg")
osm_land <- st_read(filepath)
keysss <- osm_land %>% st_drop_geometry() %>% select(land, key) %>% distinct() %>% arrange(key, land)
```


```{r}
bbox <- c(xmin = -122.5, ymin = 37.8, xmax = -122.4, ymax = 37.7)
bbox_polygon <- st_as_sfc(st_bbox(bbox))
st_crs(bbox_polygon) <- 4326
sanfran_sensors <- st_intersection(pa_sf, bbox_polygon)
sf_roads <- osm_roads %>% filter(sensor_index %in% unique(sanfran_sensors$sensor_index))
sf_land <- osm_land %>% filter(sensor_index %in% unique(sanfran_sensors$sensor_index))


# "landuse", "natural", "leisure", "boundary"


thiskey = "natural"
notthis <- osm_land %>% filter(key != thiskey)
onlythis <- osm_land %>% filter(key == thiskey, (!osm_id %in% notthis$osm_id))
# onlythis <- osm_land %>% filter(key == thiskey, (!osm_id %in% notthis$osm_id), land %in% c("national_park", "protected_area"))

# onlythis <- osm_land %>% filter(key == thiskey, land %in% c("hazard"))

mapview(onlythis)

# only landuse: 46,642
# only natural: 25703
# only leisure: 59,877
# only boundary: 247 boundary (national_park,protected_area,hazard)

x <- osm_p
x2 <- osm_mp
# 829462466
osm_p %>% filter(osm_id == "829462466")
osm_mp %>% filter(osm_id == "829462466")



landuseorleisure <- osm_land %>% filter(key %in% c("landuse", "leisure"))
onlynatural <- osm_land %>% 
  select(osm_id, land, key) %>%
  filter(key == "natural", (!osm_id %in% landuseorleisure$osm_id)) %>%
  st_make_valid() %>% 
  distinct(osm_id, .keep_all = TRUE)
onlyboundary <- osm_land %>% filter(key == "boundary", (!osm_id %in% landuseorleisure$osm_id),
                                    land %in% c("national_park", "protected_area")) %>%
  distinct(osm_id, .keep_all = TRUE)


mapview(onlynatural, col.regions = "green", alpha.regions = 0.5) + mapview(onlyboundary, col.regions = "blue", alpha.regions = 0.5)
```

Highway meanings
https://taginfo.openstreetmap.org/keys/highway#values
https://taginfo.geofabrik.de/north-america:us:california:norcal/keys/landuse#overview

Get highway colors from osm to split them


Download Land Use

natural: scrub, wood,
landuse:	residential,	industrial,	farmland
boundary:	protected_area
leisure:	park
waterway:	stream
check what to get from here
https://www.openstreetmap.org/relation/13070397#map=14/37.68559/-122.42435
https://osmlanduse.org/#10.716666666666669/-122.21682/37.73751/0/
OSM keys used for classification
Classification of the landuse/landcover classes are similar to the classification level 2 of the CORINE Landcover classes.
The following OSM keys were used to form the respective class:

1.1. Urban Fabric
residential,  garages

1.2. Industrial, commercial and transport units
railway,  industrial,  commercial,  retail,  harbour,  port,  lock,  marina

1.3. Mine, dump and construction sites
quarry,  construction,  landfill,  brownfield

1.4. Artificial, non-agricultural vegetated areas
stadium,  recreation_ground,  golf_course,  sports_center,  playground
pitch,  village_green,  allotments,  cemetery,  park,  zoo,  track,  garden,  raceway

2.1. Arable Land
greenhouse_horticulture,  greenhouse,  farmland,  farm,  farmyard

2.2. Permanent Crops
vineyard,  orchard

2.3. Pastures
meadow

3.1. Forests
forest,  wood

3.2. Shrub and/or herbaceous vegetation associations
grass,  greenfield,  scrub,  heath,  grassland

3.3. Open spaces with little or no vegetation
cliff,  fell,  sand,  scree,  beach,  mud,  glacier,  rock

4.1. Inland wetlands
marsh,  wetland

4.2. Coastal wetlands
salt_pond,  tidal

5. Water bodies
water,  riverbank,  reservoir,  basin,  dock,  canal,  pond

```{}
# # bay area
# bbox <- c(xmin = -123.8, ymin = 36.9, xmax = -121.0, ymax = 39.0)

# tiny bbox
bbox <- c(xmin = -122.46, ymin = 37.76, xmax = -122.42, ymax = 37.72)

bbox_sf <- st_as_sfc(st_bbox(bbox))
st_crs(bbox_sf) <- 4326

landuse_sf <- opq(bbox = bbox_sf) %>%
add_osm_feature(key = "landuse") %>%
osmdata_sf()

leisure_sf <- opq(bbox = bbox_sf) %>%
add_osm_feature(key = "leisure") %>%
osmdata_sf()

waterway_sf <- opq(bbox = bbox_sf) %>%
add_osm_feature(key = "waterway") %>%
osmdata_sf()

natural_sf <- opq(bbox = bbox_sf) %>%
add_osm_feature(key = "natural") %>%
osmdata_sf()

boundary_sf <- opq(bbox = bbox_sf) %>%
add_osm_feature(key = "boundary") %>%
osmdata_sf()
```

```{}
# Convert all downloaded data to sf objects for visualization
landuse_pol <- st_as_sf(landuse_sf$osm_polygons) %>% select(osm_id, landuse)
landuse_mpol <- st_as_sf(landuse_sf$osm_multipolygons) %>% select(osm_id, landuse)
# leisure_sf_sf <- st_as_sf(leisure_sf$osm_polygons)
# waterway_sf_sf <- st_as_sf(waterway_sf$osm_lines)
# natural_sf_sf <- st_as_sf(natural_sf$osm_polygons)
```


```{}
x <- as.data.frame(landuse_sf_sf)
x %>% 
group_by(landuse) %>%
summarise(count = n()) %>%
arrange(desc(count))

osmids <- x %>% filter(is.na(landuse)) %>% mutate(osm_id = as.numeric(osm_id)) %>% pull(osm_id)

# node, way, or relation
osmObj <- opq_osm_id(type = "node", id = osmids) %>%
opq_string() %>%
osmdata_sf()

osmObj$osm_points
```


```{}
colnames(x)

unique(x$landuse)

x2 <- x %>% select(where(~ sum(is.na(.)) / nrow(x) < 0.9))

library(DataOverviewR)

data_dictionary(x2)


```

```{}
img_path <- file.path("../docs", "plots", "landuse-map.png")
if (!file.exists(img_path)) {
map_path <- file.path("../docs", "maps", "landuse-map.html")
m <- leaflet() %>%
addCircleMarkers(data = pa_sf, color = "#AA44AA", weight = 1, radius = 1) %>%
addPolygons(data = purpleairs_buffers,
color = "#AA44AA", opacity = 0.5, weight = 1,
fillColor = "#AA44AA", fillOpacity = 0.2) %>%
addPolygons(data = landuse_sf_sf,
color = "#4682B4", opacity = 0.5, weight = 1,
fillColor = "#4682B4", fillOpacity = 0.2) %>%
addProviderTiles("CartoDB") %>%
setView(lng = -122.4194, lat = 37.7749, zoom = 12)
saveWidget(m, file = map_path)
webshot(map_path, file = img_path)
}

knitr::include_graphics(img_path)
```


```{}

# Plot all features together
ggplot() +
geom_sf(data = natural_sf_sf, fill = "green", alpha = 0.4, color = NA) +
geom_sf(data = landuse_sf_sf, fill = "blue", alpha = 0.4, color = NA) +
geom_sf(data = boundary_sf_sf, fill = "purple", alpha = 0.4, color = NA) +
geom_sf(data = leisure_sf_sf, fill = "yellow", alpha = 0.4, color = NA) +
geom_sf(data = waterway_sf_sf, color = "cyan", size = 0.5) +
theme_minimal() +
labs(title = "Natural and Land Use Features in San Francisco")

leaflet() %>%
addTiles() %>%
addPolygons(data = natural_sf_sf, fillColor = "#228B22", fillOpacity = 0.4, color = NA) %>%
addPolygons(data = landuse_sf_sf, fillColor = "#4682B4", fillOpacity = 0.4, color = NA) %>%
addPolygons(data = boundary_sf_sf, fillColor = "#800080", fillOpacity = 0.4, color = NA) %>%
addPolygons(data = leisure_sf_sf, fillColor = "#FFD700", fillOpacity = 0.4, color = NA) %>%
addPolylines(data = waterway_sf_sf, color = "#00FFFF", weight = 1) %>%
addProviderTiles("CartoDB.Positron") %>%
addLegend(
position = "bottomright",
colors = c("#228B22", "#4682B4", "#800080", "#FFD700", "#00FFFF"),
labels = c("Natural", "Landuse", "Boundary", "Leisure", "Waterway"),
title = "Features"
)
```


Map of OpenStreetMap Roads

```{r, plot-roads, results = FALSE}
img_path <- file.path("../docs", "plots", "roads-map.png")
if (!file.exists(img_path)) {
  bayarea_roads <- st_read(file.path("data", "raw", "bayarea_roads_osm.gpkg"), quiet = TRUE)
  map_path <- file.path("../docs", "maps", "roads-map.html")
  m <- leaflet() %>%
    addPolylines(data = bayarea_roads, 
                 color = "#DD94A1", opacity = 0.5, weight = 1) %>%
    addProviderTiles("CartoDB") %>%
    setView(lng = -122.4194, lat = 37.7749, zoom = 12)
  saveWidget(m, file = map_path)
  webshot(map_path, file = img_path)
}
knitr::include_graphics(img_path)
```

Map of OpenStreetMap Buildings

```{}
# ```{r, plot-buildings, results = FALSE}
"#D7D0CA"

bayarea_buildings <- st_read(file.path("data", "raw", "bayarea_buildings_osm.gpkg"), quiet = TRUE)
ca <- counties("California", cb = TRUE, progress_bar=FALSE)

ggplot() + 
geom_sf(data = ca, color="black", fill="antiquewhite", size=0.25) +
geom_sf(data = bayarea_buildings, color="darkorange3", fill = "darkorange") +
coord_sf(xlim = c(-123.8, -121.0), ylim = c(36.9, 39.0)) +
theme(panel.background = element_rect(fill = "aliceblue")) + 
xlab("Longitude") + ylab("Latitude") +
ggtitle("OSM Buildings")
```

Map of OpenStreetMap Trees

```{}
# ```{r, plot-trees, results = FALSE} -->
bayarea_trees <- st_read(file.path("data", "raw", "bayarea_trees_osm.gpkg"), quiet = TRUE)
ca <- counties("California", cb = TRUE, progress_bar=FALSE)

ggplot() + 
geom_sf(data = ca, color="black", fill="antiquewhite", size=0.25) +
geom_sf(data = bayarea_trees, color = "darkgreen", fill="forestgreen", size = 0.1) +
coord_sf(xlim = c(-123.8, -121.0), ylim = c(36.9, 39.0)) +
theme(panel.background = element_rect(fill = "aliceblue")) + 
xlab("Longitude") + ylab("Latitude") +
ggtitle("OSM Trees")
```



```{}
## Select small area of San Francisco to map

bbox <- c(xmin = -122.46, ymin = 37.76, xmax = -122.42, ymax = 37.72)
bbox_polygon <- st_as_sfc(st_bbox(bbox))
st_crs(bbox_polygon) <- 4326
sanfran_roads <- st_intersection(bayarea_roads, bbox_polygon)
sanfran_buildings <- st_intersection(bayarea_buildings, bbox_polygon)
sanfran_trees <- st_intersection(bayarea_trees, bbox_polygon)
sanfran_sensors <- st_intersection(filtered_sensors, bbox_polygon)
sanfran_buffers <- st_intersection(purpleairs_buffers, bbox_polygon)


## Plot san fran city
leaflet() %>%
# addPolylines(data = bbox_polygon, color = "black", weight = 3, opacity = 1, popup = "Bounding Box") %>%
addPolylines(data = sanfran_roads, color = "cornflowerblue", weight = 1, popup = "Roads") %>%
addPolygons(data = sanfran_buildings, fillOpacity = 1, weight = 1, color="darkorange", fillColor = "orange", popup = "Buildings") %>%
addCircleMarkers(data = sanfran_trees, popup = "Trees",
fillColor = "#99CC99", fillOpacity = 1,
color = "#336600", weight=1, opacity = 1, radius = 1.5) %>%
addCircleMarkers(data = sanfran_sensors, popup = "Sensors",
fillColor = "#CC66CC", fillOpacity = 1,
color = "#9933CC",weight=2, opacity = 1, radius = 5) %>%
addPolygons(data = sanfran_buffers, color = "#9933CC", weight = 1, opacity = 1, fillOpacity = 0, popup = "Buffers") %>%
addProviderTiles("CartoDB") %>% 
addLegend(colors = c("#9933CC", "cornflowerblue", "darkorange", "#336600"), 
labels = c("PurpleAir Sensors", "Roads", "Buildings", "Trees"))


## Plot san fran city


library(htmlwidgets)
library(htmltools)


major_roads <- c("motorway", "motorway_link", "trunk", "primary")

# group road types to major and minor
osm_roads_pa <- bayarea_roads %>%
mutate(highway = ifelse(is.na(highway), "NA", highway)) %>%
mutate(road_type = ifelse(highway %in% major_roads, "Major", "Minor"),
road_length = round(st_length(geom),2)) 

# Define building categories
house <- c("house", "detached", "semidetached_house", "houses", "farm")
apartments <- c("residential", "apartments")
undefined <- c("NA", "yes")

# Categorize buildings
buildings <- bayarea_buildings %>% 
mutate(building = ifelse(is.na(building), "NA", building)) %>%
mutate(building_type = case_when(
building %in% undefined ~ "Undefined",
building %in% house ~ "House",
building %in% apartments ~ "Apartments",
TRUE ~ "Other"
),
building_area = round(st_area(geom),2))

# Remove units from road_length
# attributes(road_lengths$road_length) = NULL

# Define colors for road types
road_colors <- colorFactor(c("cornflowerblue", "lightblue"), domain = c("Major", "Minor"))
building_colors <- colorFactor(c("orangered", "darkorange", "peachpuff", "papayawhip"), 
domain = c("Other", "Apartments", "House", "Undefined"))

m <- leaflet() %>%
addPolylines(data = osm_roads_pa, color = ~road_colors(road_type), weight = 2,
label = ~HTML(paste(htmlEscape(road_type), " (", htmlEscape(highway),
")<br>Length: ", htmlEscape(road_length), " [m]")),
popup = "Roads") %>%
addPolygons(data = buildings, fillOpacity = 1, weight = 1,
color=~building_colors(building_type), fillColor = ~building_colors(building_type),
label = ~HTML(paste(htmlEscape(building_type), " (", htmlEscape(building),
")<br>Area: ", htmlEscape(building_area), " [m^2]"),
popup = "Buildings")) %>%
addCircleMarkers(data = bayarea_trees, popup = "Trees", label = "Tree",
fillColor = "#99CC99", fillOpacity = 1,
color = "#336600", weight=1, opacity = 1, radius = 2) %>%
addCircleMarkers(data = z, popup = "Sensors",
label = ~paste("Sensor:", htmlEscape(sensor_index)),
fillColor = "#CC66CC", fillOpacity = 1, color = "#9933CC",
weight=2, opacity = 1, radius = 5) %>%
addPolylines(data = purpleairs_buffers, color = "#9933CC", weight = 1,
opacity = 1, popup = "Buffers") %>%
addProviderTiles("CartoDB") %>% 
addLegend(colors = c("#9933CC", "#336600", "cornflowerblue", "lightblue",
"orangered", "darkorange", "peachpuff", "oldlace"), 
labels = c("PurpleAir Sensors", "Trees", "Major Roads", "Minor Roads",
"Other", "Apartments", "House", "Undefined"))
m
saveWidget(m, file="san_fran_map.html")
```
