---
title: "Uber Speeds data"
output: github_document
knit: (function(input, ...) {rmarkdown::render(input, output_dir = "../docs", )})
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(fig.path = "../docs/plots/")
```

Load required libraries

```{r, load-libraries, message = FALSE, warning = FALSE}
library(dplyr)         # Data manipulation
library(data.table)    # Working with large files
library(DataOverviewR) # Data dictionary and summary
library(ggplot2)
library(here)
```

```{r, read-uber1, echo = FALSE}
uber201801 <- fread(here("code", "data", "raw", "Uber", "Speeds",
                              "movement-speeds-hourly-san-francisco-2018-1.csv"))
```

---

**Data Dictionary**

```{r, speeds-data-dict, echo = FALSE}
# utc_timestamp - Date & time of observations in UTC format
# segment_id - Special Ids assigned to road segments by Uber
# start_junction_id - Junction where the segment starts i.e., a roundabout
# end_junction_id - Junction where the segment ends
# speed_mph_mean - Mean speed of vehicles within an hour
# speed_mph_stddev - Standard deviation of speed of vehicles in an hour
# The above columns can uniquely identify a street segment and its speed. However, since the segments and junctions correspond to Uber's internal identification models. There also exists Open Street Map (OSM) Ids, that correspond to OSM Way Id and OSM Node Id respectively.
# 
# OSM Way and Node Ids define roads and nodes connecting them. Uber has its own implementation of road structures because speeds can vary a lot within a single OSM Way Id. That is why multiple Segment Ids correspond to a single OSM Way Id. Fortunately, OSM Ids are also provided in the same table
# 
# osm_way_id - OSM Way Id with One to Many relationship with segment_id
# osm_start_node_id - Start node Id of OSM corresponding to start_junction_id
# osm_end_node_id - End node Id of OSM corresponding to end_junction_id

# The datasets contain hourly speed observations of three months. In Open Street Map (OSM), a way defines a line connecting two nodes e.g., a street or a road. Whereas OSM nodes are the points in real world such as a junction. For the speed analysis, a street can be uniquely identified via osm_way_id, osm_start_node_id, and osm_end_node_id. Note that we need all three parameters to identify a street speed because a street can be one way or two way. In case of later, osm-way-id will be the same for both ways but their node ids would swap.
# 
# Mean Speeds in miles per hour are provided per hour. But in the dataset, their are missing values and data is not available for each hour.
desc <- data_description(uber201801, var_desc = c(
  "year" = "Year.",
  "month" = "Month.",
  "day" = "Day.",
  "hour" = "Hour in 24-hour format.",
  "utc_timestamp" = "Timestamp in Coordinated Universal Time (UTC)",
  "segment_id" = "Unique identifier for the road segment on which the speed is measured.",
  "start_junction_id" = "Unique identifier for the starting junction (node) of the road segment.",
  "end_junction_id" = "Unique identifier for the ending junction (node) of the road segment.",
  "osm_way_id" = "OpenStreetMap (OSM) ID for the road segment or way.",
  "osm_start_node_id" = "OpenStreetMap (OSM) ID for the starting node of the segment.",
  "osm_end_node_id" = "OpenStreetMap (OSM) ID for the ending node of the segment.",
  "speed_mph_mean" = "Mean speed in miles per hour recorded on the segment during the observation period.",
  "speed_mph_stddev" = "Standard deviation of the speed in miles per hour recorded on the segment during the observation period."
))
data_dictionary(uber201801, data_title = "Uber Movement Speeds - Jan 2018", descriptions = desc, hide = c("top_n", "NA_Percentage", "NA_Count", "n_unique"))
```

**View data**

```{r, view-speeds1, echo = FALSE}
knitr::kable(head(uber201801, 3),
             row.names = FALSE,
             format = "markdown")

# remove to free up memory
rm(uber201801)
```

---

### Processing Uber roads

Write Uber 2018 to monthly files
```{r, eval = FALSE}
# Set chunk size
chunk_size <- 10000000

# Process file in chunks
offset <- 0

while(FALSE) {
  # Read chunk
  chunk <- fread(here("code", "data", "raw", "uber_2018.csv"), 
                 skip = offset, 
                 nrows = chunk_size)
  names(chunk) <- c("utc_timestamp", "osm_way_id", "speed_mph_mean")
  
  # Break if no more data
  if(nrow(chunk) == 0) break
  
  # Extract month without converting to date (more memory efficient)
  chunk[, month := as.numeric(substr(utc_timestamp, 6, 7))]
  
  # Append each month's data to respective files
  for(m in 1:12) {
    month_data <- chunk[month == m]
    if(nrow(month_data) > 0) {
      # Remove month column before writing
      month_data[, month := NULL]
      fwrite(month_data, 
             here("code", "data", "processed", "Uber", sprintf("uber2018_%02d.csv", m)),
             append = TRUE)
    }
  }
  
  # Update offset and print progress
  offset <- offset + chunk_size
  cat(sprintf("Processed %d rows...\n", offset))
  
  # Clear memory
  rm(chunk, month_data)
  gc()
}
```

Write Uber 2019 to monthly files
```{r, eval = FALSE}
# Set chunk size
chunk_size <- 10000000

# Process file in chunks
offset <- 0

while(FALSE) {
  # Read chunk
  chunk <- fread(here("code", "data", "raw", "uber_2019.csv"), 
                 skip = offset, 
                 nrows = chunk_size)
  names(chunk) <- c("utc_timestamp", "osm_way_id", "speed_mph_mean")
  
  # Break if no more data
  if(nrow(chunk) == 0) break
  
  # Extract month without converting to date (more memory efficient)
  chunk[, month := as.numeric(substr(utc_timestamp, 6, 7))]
  
  # Append each month's data to respective files
  for(m in 1:12) {
    month_data <- chunk[month == m]
    if(nrow(month_data) > 0) {
      # Remove month column before writing
      month_data[, month := NULL]
      fwrite(month_data, 
             here("code", "data", "processed", "Uber", sprintf("uber2019_%02d.csv", m)),
             append = TRUE)
    }
  }
  
  # Update offset and print progress
  offset <- offset + chunk_size
  cat(sprintf("Processed %d rows...\n", offset))
  
  # Clear memory
  rm(chunk, month_data)
  gc()
}
```

Read file: Unique roads surrounding purpleair sensors
```{r, read-osmroads}
# osm roads surrounding purple air sensors
# columns: sensor_index, osm_id, highway
osm_roads <- fread(here("code", "data", "processed", "osm_roads.csv"))
unique_roads <- unique(osm_roads$osm_id)
```

Free flow speeds per road
```{r, freeflow, eval = FALSE}
# Initialize an empty data table to store the top 5% rows
top_5_percent <- data.table()

# Iterate through all months
for (m in 1:12) {
  # Read the monthly file (columns: utc_timestamp, osm_way_id, speed_mph_mean)
  monthly_data <- fread(
    here("code", "data", "processed", "Uber",
              paste0("uber2018_", sprintf("%02d", m), ".csv")))
  
  # Filter to keep only roads in your unique_roads dataset
  monthly_data <- monthly_data[osm_way_id %in% unique_roads]
  
  # Calculate the threshold for the top 5% of speeds
  monthly_top_5 <- monthly_data[, .SD[order(-speed_mph_mean)][1:ceiling(0.05 * .N)], by = osm_way_id]
  
  # Append the top 5% rows to the result data table
  top_5_percent <- rbind(top_5_percent, monthly_top_5, fill = TRUE)
  
  # Remove the current month's data to save memory
  rm(monthly_data)
  gc()  # Trigger garbage collection
}

# Calculate free flow speed (95th percentile) for each osm_way_id
top_5_percent[, speed_mph_mean := as.numeric(speed_mph_mean)]

free_flow_speed <- top_5_percent[, .(free_flow_speed = round(quantile(speed_mph_mean, 0.95, na.rm = TRUE))), by = osm_way_id]

# Save the result
fwrite(free_flow_speed, here("code", "data", "processed", "Uber", "free_flow_speeds.csv"))
```

```{r, read-freeflow, include = FALSE}
# columns: osm_way_id	free_flow_speed
free_flow_speed <- fread(here("code", "data", "processed", "Uber", "free_flow_speeds.csv"))
```

Group OSM road types using free flow speeds of highway types

```{r, boxplot-data}
# drop sensor_index (causes duplicate roads)
roads_data <- osm_roads %>% 
  select(osm_id, highway) %>% 
  distinct() %>%
  mutate(osm_id = as.numeric(osm_id))

roads_data <- left_join(roads_data, free_flow_speed, by = c("osm_id" = "osm_way_id"))

roads_analysis <- roads_data %>% 
  select(osm_id, highway, free_flow_speed) %>% 
  distinct() %>% 
  group_by(highway) %>% 
  filter(!is.na(free_flow_speed)) %>%
  summarise(
    min_f = min(free_flow_speed, na.rm = TRUE),
    f25 = quantile(free_flow_speed, 0.25, na.rm = TRUE),
    median_f = median(free_flow_speed, na.rm = TRUE),
    f75 = quantile(free_flow_speed, 0.75, na.rm = TRUE),
    max_f = max(free_flow_speed, na.rm = TRUE),
    countf = n()
  ) %>%
  mutate(road_type = case_when(
    highway %in% c("motorway", "motorway_link", "trunk", "trunk_link") ~ "major",
    highway %in% c("primary", "primary_link", "secondary", "tertiary") ~ "moderate",
    highway %in% c("residential", "unclassified", "service", "living_street",
                   "secondary_link", "tertiary_link") ~ "minor",
    TRUE ~ "minor"
  )) %>% arrange(road_type, desc(countf))
```

Box plot to decide on road type grouping

```{r, boxplot}
roads_analysis <- roads_analysis %>%
  mutate(road_type = factor(road_type, levels = c("major", "moderate", "minor")))
# Boxplot to decide on road type grouping (using free flow speeds)
ggplot(roads_analysis, aes(x = reorder(highway, median_f), y = median_f)) +
  geom_boxplot(
    aes(ymin = min_f, lower = f25, middle = median_f, upper = f75,
        ymax = max_f, fill = road_type), stat = "identity") +
  geom_point(aes(size = countf), color = "black", alpha = 0.6) +
  # scale_fill_brewer(palette = "YlOrRd") +
  scale_fill_viridis_d(option = "C",begin=0.2, end=0.9) +
  scale_size_continuous(name = "Count of roads") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Free Flow Speed Distribution by Highway Type",
       x = "Highway Type", y = "Free Flow Speed (mph)", fill = "Road Type")
```


```{r, read-files, include = FALSE}
# columns: osm_way_id,	free_flow_speed
free_flow_speed <- fread(here("code", "data", "processed", "Uber", "free_flow_speeds.csv"))
# osm roads surrounding purple air sensors
# columns: sensor_index, osm_id, highway
osm_roads <- fread(here("code", "data", "processed", "osm_roads.csv"))
unique_roads <- unique(osm_roads$osm_id)
```

Calculate hourly congestion & speeds for each purpleair sensor & road type

```{r, process-speeds, eval = FALSE}
if (!file.exists(here("code", "data", "processed", "speeds_data.csv"))) {
  
  all_months <- list()
  counter <- 1
  for (y in c("2018_", "2019_")) {
    for (m in 1:12) {
      # Read the monthly file (columns: utc_timestamp, osm_way_id, speed_mph_mean)
      monthly_data <- fread(here("code", "data", "processed", "Uber", paste0(
        "uber", y, sprintf("%02d", m), ".csv")))
      
      # Filter to keep only roads surrounding purple air sensors
      monthly_data <- monthly_data[osm_way_id %in% unique_roads]
      
      # join to get free_flow_speed column
      monthly_data <- left_join(monthly_data, free_flow_speed, by = c("osm_way_id"))
      
      # join to get sensor_index & highway columns
      monthly_data <- left_join(monthly_data, osm_roads, by = c("osm_way_id" = "osm_id"))
      
      # categorize roads by type
      monthly_data <- monthly_data %>% 
        mutate(road_type = case_when(
          highway %in% c("motorway", "motorway_link", "trunk", "trunk_link") ~ "major",
          highway %in% c("primary", "primary_link", "secondary", "tertiary") ~ "moderate",
          highway %in% c("residential", "unclassified", "service", "living_street",
                         "secondary_link", "tertiary_link") ~ "minor",
          TRUE ~ "minor"
        ))
      
      # congestion and speed aggregated by sensor and road type (hourly)
      monthly_agg <- monthly_data %>% 
        mutate(congestion_ratio = speed_mph_mean / free_flow_speed) %>% 
        group_by(utc_timestamp, sensor_index, road_type) %>%
        summarize(
          mean_congestion = round(mean(congestion_ratio, na.rm = TRUE), 3),
          mean_speed = round(mean(speed_mph_mean, na.rm = TRUE), 1),
          n_measurements = n()
        ) %>%
        ungroup()
      
      # Pivot wider to get one row per sensor-hour
      monthly_wide <- monthly_agg %>%
        pivot_wider(
          id_cols = c(utc_timestamp, sensor_index),
          names_from = road_type,
          values_from = c(mean_congestion, mean_speed, n_measurements),
          values_fill = list(mean_congestion = NA, mean_speed = NA, n_measurements = 0)
        )
      
      # Store in list
      all_months[[counter]] <- monthly_wide
      counter <- counter + 1
    }
  }
  # Combine all months
  speeds_data <- rbindlist(all_months, fill = TRUE)
  
  # Save final dataset
  fwrite(speeds_data, here("code", "data", "processed", "speeds_data.csv"))
}
```

```{r, include = FALSE}
speeds_data <- fread(here("code", "data", "processed", "speeds_data.csv"))
```

---

**Data Dictionary**

```{r, speeds-data-dict1, echo = FALSE}
desc <- data_description(speeds_data, var_desc = c(
  "utc_timestamp" = "Timestamp in Coordinated Universal Time (UTC).",
  "sensor_index" = "PurpleAir sensor unique identifier.",
  "mean_congestion_major" = "Average congestion ratio on major roads (e.g., motorways, trunk).",
  "mean_congestion_moderate" = "Average congestion ratio on moderate roads (e.g., primary, secondary, tertiary).",
  "mean_congestion_minor" = "Average congestion ratio on minor roads (e.g., residential, unclassified).",
  "mean_speed_major" = "Average speed (mph) on major roads.",
  "mean_speed_moderate" = "Average speed (mph) on moderate roads.",
  "mean_speed_minor" = "Average speed (mph) on minor roads.",
  "n_measurements_major" = "Number of measurements for major roads.",
  "n_measurements_moderate" = "Number of measurements for moderate roads.",
  "n_measurements_minor" = "Number of measurements for minor roads."
))
data_dictionary(speeds_data, data_title = "Processed Uber Movement Speeds", descriptions = desc, hide = c("top_n", "NA_Percentage", "NA_Count", "n_unique"))

# Show missing value patterns
data_dictionary(speeds_data, 
                data_title = "Missing Values Summary",
                hide = c("top_n", "Type", "N_Unique"))
```

**View data**

```{r, view-speeds, echo = FALSE}
knitr::kable(head(speeds_data, 3),
             row.names = FALSE,
             format = "markdown")
```

---
