---
title: "Exploratory Data Analysis: PurpleAir"
output: github_document
knit: (function(input, ...) {rmarkdown::render(input, output_dir = "../docs", )})
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(fig.path = "../docs/plots/")
```

```{r, load-libraries, message = FALSE, warning = FALSE, echo = FALSE}
library(dplyr)         # For data manipulation
library(data.table)    # Working with large files
library(ggplot2)       # Plots
library(plotly)        # Interactive plots
library(lubridate)     # Dates
library(sf)            # Shapefiles
library(leaflet)       # Interactive maps
library(kableExtra)    # Printing formatted tables
library(zoo)           # for rolling calculations
library(purpleAirAPI)  # Download PurpleAir Data
library(tidyr)         # Reshape data
library(DataOverviewR) # Data dictionary and summary
library(reshape2)      # melt correlations
library(tidyverse)
library(timeDate)      # holidays
```

```{r, read-purpleair-csv, echo = FALSE, results = "hide", message = FALSE}
# Read purple air data
purpleair_data <- fread(file.path("data", "raw", "purpleair_2018-01-01_2019-12-31.csv")) %>% distinct()
purpleair_data <- purpleair_data %>% 
  select(time_stamp, rssi, uptime, memory, humidity, temperature, pressure,
         analog_input, pm2.5_alt, pm2.5_alt_a, pm2.5_alt_b, sensor_index, location_type)

aqs_data <- fread(file.path("data", "raw", "EPA_airquality.csv"))

total_rows <- nrow(purpleair_data)
```

---

**Data Summary**

```{r, pa-summ, echo = FALSE}
data_summary(purpleair_data, data_title = "Purple Air Summary")
```

---

Comparing PurpleAir air quality measurements with AQS

```{r, pa-aqs-quantiles, echo = FALSE}
quants_vals <- c(0.75, 0.95, 0.99, 0.999, 1)
quants <- c("75%","95%","99%","99.9%", "max")
summary_statistics <- purpleair_data %>%
  reframe(
    across(c(pm2.5_alt, pm2.5_alt_a, pm2.5_alt_b),
           ~tibble(val = round(quantile(.x, quants_vals, na.rm = TRUE),0)),
           .unpack = TRUE)) %>%
  mutate(`Quantiles` = quants,
         aqs = quantile(aqs_data$pm25, quants_vals, na.rm = TRUE)) %>%
  rename(`AQS PM2.5` = aqs,
         `PurpleAir ALT` = pm2.5_alt_val,
         `Channel A ALT` = pm2.5_alt_a_val,
         `Channel B ALT` = pm2.5_alt_b_val
  ) %>%
  select(`Quantiles`, `AQS PM2.5`, everything())

summary_statistics_long <- summary_statistics |> 
  pivot_longer(`AQS PM2.5`:`Channel B ALT`) |> 
  pivot_wider(names_from = Quantiles, values_from = value)

summary_statistics_long <- summary_statistics_long %>% rename(Quantiles = name)

knitr::kable(summary_statistics_long,
             row.names = FALSE,
             format = "markdown")
```

Box Plots

```{r, box-plot, echo = FALSE}
img_path <- file.path("../docs", "plots", "boxplots.png")

if (!file.exists(img_path)) {
  # Combine relevant columns into a single data frame
  combined_data <- data.frame(
    ALT = purpleair_data$pm2.5_alt,
    ALT_A = purpleair_data$pm2.5_alt_a,
    ALT_B = purpleair_data$pm2.5_alt_b
  )
  
  # Reshape the data to long format for plotting
  long_data <- combined_data %>%
    pivot_longer(cols = everything(),  # All columns
                 names_to = "Measurement", 
                 values_to = "Value")
  
  # Add AQS data
  aqs <- data.frame(Measurement = "AQS", Value = aqs_data$pm25)
  long_data <- rbind(long_data, aqs)
  long_data <- long_data %>% na.omit()
  
  # Box plots
  p <- ggplot(long_data, aes(x = Measurement, y = Value)) +
    geom_boxplot(fill = "lightblue") +
    # ylim(0, 50) +
    coord_cartesian(ylim = c(0, 30)) +
    labs(title = "Boxplot Comparison: PurpleAir and AQS PM2.5 Data",
         subtitle = "y-axis limit set to 30; more data points beyond the limit",
         x = "",
         y = "PM2.5 Value (µg/m³)") +
    theme_minimal()
  ggsave(filename = img_path, plot = p, width = 6, height = 4)
}
knitr::include_graphics(img_path)
```

Plot Sensors by Month

```{r, monthly-sensors, echo = FALSE}
img_path <- file.path("../docs", "plots", "sensors-by-month.png")

if (!file.exists(img_path)) {
  # Add column for month
  purpleair_data$month <- format(as.Date(purpleair_data$time_stamp), "%Y-%m")
  
  # Sensors for each month
  monthly_sensors <- purpleair_data %>% select(month, sensor_index) %>% distinct()
  
  sensor_counts <- monthly_sensors %>%
    group_by(month) %>%
    summarise(sensor_count = n_distinct(sensor_index))
  
  p <- ggplot(sensor_counts, aes(x = month, y = sensor_count)) +
    geom_bar(stat = "identity", fill = "lavender", color = "black") +
    labs(title = "Active PurpleAir Sensors By Month",
         x = "Month",
         y = "Number of Sensors") +
    scale_y_continuous(breaks = seq(0, max(sensor_counts$sensor_count) + 100, by = 100)) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  ggsave(filename = img_path, plot = p, width = 6, height = 4)
  
}
knitr::include_graphics(img_path)
```

Create temporal features

```{r, temporal-features}
# Get holidays for 2018 and 2019
holidays <- as.Date(c(holidayNYSE(2019), holidayNYSE(2018)))

purpleair_data <- purpleair_data %>%
  mutate(
    time_stamp = lubridate::as_datetime(time_stamp),
    local_timestamp = with_tz(time_stamp, tzone = "America/Los_Angeles"),
    local_date = as.Date(local_timestamp, tz="America/Los_Angeles"),
    dow = lubridate::wday(local_timestamp),
    hour = lubridate::hour(local_timestamp),
    day = lubridate::day(local_timestamp),
    month = lubridate::month(local_timestamp),
    year = lubridate::year(local_timestamp),
    holiday = ifelse(local_date %in% holidays, 1, 0)
  ) %>% select(-local_timestamp, -local_date)
```

Histograms for numerical columns

```{r, hist, echo = FALSE}
cols <- c("pm2.5_alt", "pm2.5_alt_a", "pm2.5_alt_b", "rssi", "uptime", 
          "memory", "humidity", "temperature", "pressure", "analog_input")

# Loop through the columns and plot histograms
for (col in cols) {
  x <- purpleair_data[[col]]
  
  # If max is much larger than the 99th percentile, filter the data for plotting
  p99 <- quantile(x, 0.99, na.rm = TRUE)
  if (max(x, na.rm = TRUE) > 2 * p99) {
    x <- x[x >= quantile(x, 0.01, na.rm = TRUE) & x <= p99]
  }
  hist(x, main = paste("Histogram of", col))
}
```

Bar Plots for categorical columns

```{r, bar-plots, echo = FALSE}
cols <- c("dow", "hour", "day", "month", "holiday", "year", "location_type")
for (col in cols) {
  # Plot the histogram using barplot for categorical data
  barplot(table(purpleair_data[[col]]), main = paste("Histogram of", col), ylab = "Frequency")
}
```

ALT: Channel A vs B

```{r, plot-channel-a-b-alt, warning=FALSE, echo = FALSE}
img_path <- file.path("../docs", "plots", "purpleair_alt_avsb.png")

if (!file.exists(img_path)) {
  purpleair_temp <- copy(purpleair_data) 
  purpleair_temp$pm2.5_alt_a[is.na(purpleair_temp$pm2.5_alt_a)] <- -1
  purpleair_temp$pm2.5_alt_b[is.na(purpleair_temp$pm2.5_alt_b)] <- -1
  purpleair_temp$missingchannel <- ifelse(purpleair_temp$pm2.5_alt_a == -1 | purpleair_temp$pm2.5_alt_b == -1, "Missing", "Valid")
  
  p <- ggplot(purpleair_temp, aes(x = pm2.5_alt_a, y = pm2.5_alt_b, color = missingchannel)) +
    geom_point() +
    scale_color_manual(values = c("Valid" = "black", "Missing" = "red"), name = NULL) +
    labs(x = "Channel A PM2.5 (µg/m³)",
         y = "Channel B PM2.5 (µg/m³)",
         title = "ALT: PM2.5 Channel A vs B") +
    theme_minimal()
  
  ggsave(filename = img_path, plot = p, width = 6, height = 4)
}

knitr::include_graphics(img_path)
```

Check associations between variables

```{r, cor-heatmap, echo = FALSE}
# Correlation heatmap
img_path <- file.path("../docs", "plots", "purpleair-cor.png")

if (!file.exists(img_path)) {
  cor_vars <- purpleair_data %>% 
    select_if(is.numeric) %>% select(pm2.5_alt, everything(), -pm2.5_alt_a, -pm2.5_alt_b)
  
  correlation_matrix <- cor(cor_vars, use = "complete.obs")
  
  melted_correlation <- melt(correlation_matrix)
  
  p <- ggplot(data = melted_correlation, aes(Var1, Var2, fill = value)) + 
    geom_tile() + 
    scale_fill_gradient2(low = "red", high = "green", mid = "white",
                         midpoint = 0, limit = c(-1, 1), name = "Correlation") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
  ggsave(filename = img_path, plot = p, width = 6, height = 4)
}

knitr::include_graphics(img_path)
```

```{r, scatterplots, echo = FALSE}
img_path <- file.path("../docs", "plots", "purpleair-scatter-pm25-humidity.png")

if (!file.exists(img_path)) {
  x <- purpleair_data$pm2.5_alt
  y <- purpleair_data$humidity
  
  colors <- ifelse(purpleair_data$location_type == 1, 
                   rgb(173, 216, 230, maxColorValue = 255, alpha = 127),
                   rgb(144, 238, 144, maxColorValue = 255, alpha = 127))
  
  png(img_path, width = 800, height = 600)
  
  plot(purpleair_data$humidity, purpleair_data$pm2.5_alt,
       col = colors,
       pch = 19,
       xlab = "Humidity",
       ylab = "PM2.5 (alt)", 
       main = "Humidity vs PM2.5 by Location Type")
  
  legend("topright", legend = c("Indoor", "Outdoor"), fill = c("lightblue", "lightgreen"))
  
  dev.off()
}

knitr::include_graphics(img_path)
```

## Remove sensor 20349 (code commented below)
```{r, remove-sensor}
purpleair_data <- purpleair_data %>%
  filter(sensor_index != 20349)
```

# remove sensors with too little readings

```{r}
timestamp_rows <- purpleair_data %>% select(time_stamp) %>% unique() %>% count() %>% pull()

sensor_counts <- purpleair_data %>%
  group_by(sensor_index) %>%
  summarise(p = round(n()/timestamp_rows,2),
            counts = n()) %>% arrange(desc(p))

# View the results
print(sensor_counts)

keep_sensors <- sensor_counts %>% filter(counts > 1500) %>% pull(sensor_index)
purpleair_data <- purpleair_data %>% filter(sensor_index %in% keep_sensors)
```


## Complete time series within each sensor date range

```{r}
# Step 1: Find the start and end time for each sensor
sensor_periods <- purpleair_data[, .(start_time = min(time_stamp, na.rm = TRUE), 
                                     end_time = max(time_stamp, na.rm = TRUE)), 
                                 by = sensor_index]

# Step 2: Create a full sequence of timestamps within the active range for each sensor
all_timestamps <- sensor_periods[, .(
  time_stamp = seq(from = start_time, to = end_time, by = "hour")
), by = sensor_index]

# Step 3: Perform a left join to include all timestamps within the active range, keeping existing data
purpleair_complete <- merge(all_timestamps, purpleair_data, by = c("sensor_index", "time_stamp"), all.x = TRUE)
```

```{r}
# Step 3: Calculate and view NA percentages for each column to verify imputation
na_percentage <- purpleair_complete[, lapply(.SD, function(x) round(mean(is.na(x)) * 100))]
na_percentage <- melt(na_percentage, variable.name = "Column", value.name = "NA_Percentage")
print(na_percentage)
```

```{r}
setDT(purpleair_complete)  # Ensure it’s a data.table

# Most frequent value (Mode)
Mode <- function(x) {
  ux <- unique(x[!is.na(x)])  # Exclude NA values
  ux[which.max(tabulate(match(x, ux)))]
}

location_type_modes <- purpleair_complete[, .(location_type_mode = Mode(location_type)), by = sensor_index]
# Join the mode values back into `purpleair_complete` using a left join
purpleair_complete_temp <- merge(purpleair_complete, location_type_modes, by = "sensor_index", all.x = TRUE)

# Use coalesce to fill `location_type` with `location_type_mode` where it's missing
purpleair_complete_temp[, location_type := coalesce(location_type, location_type_mode)]
```

```{r}
# check location type doesnt have NAs after filling with mode
na_percentage <- purpleair_complete_temp[, lapply(.SD, function(x) round(mean(is.na(x)) * 100))]
na_percentage <- melt(na_percentage, variable.name = "Column", value.name = "NA_Percentage")
print(na_percentage)
```

```{r}
# Step 1: Convert time-related features and handle `location_type` imputation
purpleair_processed_dates <- purpleair_complete_temp %>%
  group_by(sensor_index) %>%
  mutate(
    time_stamp = lubridate::as_datetime(time_stamp),
    local_timestamp = with_tz(time_stamp, tzone = "America/Los_Angeles"),
    local_date = as.Date(local_timestamp, tz = "America/Los_Angeles"),
    dow = lubridate::wday(local_timestamp),
    hour = lubridate::hour(local_timestamp),
    day = lubridate::day(local_timestamp),
    month = lubridate::month(local_timestamp),
    year = lubridate::year(local_timestamp),
    holiday = ifelse(local_date %in% holidays, 1, 0),
  ) %>%
  ungroup() %>%
  select(-pm2.5_alt_a, -pm2.5_alt_b, -local_timestamp, -local_date)
```

```{r}
setDT(purpleair_processed_dates)  # Ensure it’s a data.table

# List of columns to impute
cols_to_impute <- c("pm2.5_alt", "pressure", "analog_input", "location_type", "temperature", "humidity", "memory", "uptime", "rssi")

# Step 1: Fill within each sensor using forward fill, backward fill, and mean within each group
purpleair_processed <- purpleair_processed_dates[, (cols_to_impute) := lapply(.SD, function(x) {
    x <- na.locf(x, na.rm = FALSE)               # Forward fill
    x <- na.locf(x, fromLast = TRUE, na.rm = FALSE) # Backward fill
    x[is.na(x)] <- mean(x, na.rm = TRUE)         # Mean within each sensor
    return(x)
  }), by = sensor_index, .SDcols = cols_to_impute]

# Step 2: For any remaining NA values, fill with the global mean across all sensors
purpleair_processed[, (cols_to_impute) := lapply(.SD, function(x) {
    x[is.na(x)] <- mean(x, na.rm = TRUE)         # Global mean imputation
    return(x)
  }), .SDcols = cols_to_impute]

# Check NA percentages after imputation
na_percentage <- purpleair_processed[, lapply(.SD, function(x) mean(is.na(x)) * 100)]
print(na_percentage)
```
```{r}
# Ensure `purpleair_complete` is a data.table
setDT(purpleair_processed)

# Add lagged columns for pm2.5_alt at 6, 12, and 24-hour intervals, grouped by sensor_index
purpleair_processed_lags <- purpleair_processed[
  order(sensor_index, time_stamp),  # Ensure correct order for lag calculation
  `:=`(
    pm2.5_alt_lag1 = shift(pm2.5_alt, n = 1, type = "lag"),
    pm2.5_alt_lag6 = shift(pm2.5_alt, n = 6, type = "lag"),
    pm2.5_alt_lag12 = shift(pm2.5_alt, n = 12, type = "lag"),
    pm2.5_alt_lag24 = shift(pm2.5_alt, n = 24, type = "lag")
  ),
  by = sensor_index  # Apply within each sensor group
]

purpleair_final <- purpleair_processed_lags[!is.na(pm2.5_alt_lag6) & 
                                    !is.na(pm2.5_alt_lag12) & 
                                    !is.na(pm2.5_alt_lag24)]
```


```{r}
# Step 4: Save the processed data
write.csv(purpleair_final, file = file.path("data", "processed", "purpleair_processed_2.csv"), row.names = FALSE)
```

```{r}
library(stats)
calc_autocorrelation <- function(sensor_data) {
  acf(sensor_data$pm2.5_alt, plot = FALSE)$acf
}

sensor_autocorrelation <- purpleair_data %>%
  filter(!is.na(pm2.5_alt)) %>%
  group_by(sensor_index) %>%
  do(autocorr = calc_autocorrelation(.)) %>%
  ungroup()
```

```{r}
# Select a sensor to analyze
sensor_index <- 767

# Get the PM2.5 data for the selected sensor
sensor_data <- purpleair_data %>% 
  filter(sensor_index == 767 & !is.na(pm2.5_alt))

# Calculate the autocorrelation
autocorr <- calc_autocorrelation(sensor_data)

# Plot the autocorrelation
plot(autocorr, type = "b", main = paste("Autocorrelation for Sensor", sensor_index))
```

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)

# Update the calc_autocorrelation function
calc_autocorrelation <- function(sensor_data) {
  acf(sensor_data$pm2.5_alt, plot = FALSE)$acf[1:24]  # Return the first 24 lags
}

# Calculate autocorrelation for each sensor
sensor_autocorrelation <- purpleair_data %>%
  group_by(sensor_index) %>%
  filter(!is.na(pm2.5_alt)) %>%
  summarize(autocorr = list(calc_autocorrelation(pick(pm2.5_alt)))) %>%  # Replace cur_data() with pick()
  ungroup()

# Expand autocorrelation into long format
sensor_autocorrelation_long <- sensor_autocorrelation %>%
  unnest(cols = c(autocorr)) %>%
  group_by(sensor_index) %>%
  mutate(lag = row_number()) %>%
  ungroup()

# Summarize autocorrelation across sensors
autocorr_summary <- sensor_autocorrelation_long %>%
  group_by(lag) %>%
  summarize(
    mean_autocorr = mean(autocorr, na.rm = TRUE),
    median_autocorr = median(autocorr, na.rm = TRUE)
  )

# Plot the autocorrelation summary
ggplot(autocorr_summary, aes(x = lag)) +
  geom_line(aes(y = mean_autocorr), color = "blue") +
  geom_line(aes(y = median_autocorr), color = "red") +
  labs(x = "Lag", y = "Autocorrelation", title = "Summary of Autocorrelation Across Sensors")
```

	•	Based on the plot, you might include lags 1, 6, 12, and 24, as they represent meaningful intervals (e.g., hourly, 6-hourly, 12-hourly, daily).






<!-- ## Remove 24 hour periods of zeros or missing data -->

<!-- ```{r, rolling-zeros-missing} -->
<!-- start_time <- min(purpleair_data$time_stamp, na.rm = TRUE) -->
<!-- end_time <- max(purpleair_data$time_stamp, na.rm = TRUE) -->
<!-- all_timestamps <- seq(from = start_time, to = end_time, by = "hour") -->
<!-- all_sensor_indices <- unique(purpleair_data$sensor_index) -->
<!-- complete_timestamps <- expand.grid(sensor_index = all_sensor_indices, -->
<!--                                    time_stamp = all_timestamps) -->
<!-- purpleair_complete <- complete_timestamps %>% -->
<!--   left_join(purpleair_data, by = c("sensor_index", "time_stamp")) -->
<!-- ``` -->


<!-- ```{r, rolling-zeros-missing} -->
<!-- pa_complete <- purpleair_complete %>% -->
<!--   mutate(pm2.5_alt = ifelse(is.na(pm2.5_alt), -1, pm2.5_alt)) %>% -->
<!--   group_by(sensor_index) %>% -->
<!--   mutate( -->
<!--     is_zero_or_missing = ifelse(pm2.5_alt == 0 | pm2.5_alt == -1, 1, 0), -->
<!--     is_normal = ifelse(pm2.5_alt != 0 & pm2.5_alt != -1, 1, 0), -->
<!--     rolling_zeros_missing = rollapply(is_zero_or_missing, width = 24, FUN = sum, -->
<!--                                       align = "right", fill = NA), -->
<!--     rolling_normals = rollapply(is_normal, width = 24, FUN = sum, align = "right", fill = NA) -->
<!--   ) %>% -->
<!--   ungroup() -->

<!-- # threshold 0.8 means >=20 hours missing from a 24 hour time period -->
<!-- pa_complete <- pa_complete %>% -->
<!--   mutate( -->
<!--     proportion_zeros_missing = rolling_zeros_missing / 24, -->
<!--     flag_high_proportion = ifelse(proportion_zeros_missing >= 0.8, 1, 0) -->
<!--   ) -->

<!-- # pa_complete <- pa_complete %>% filter(flag_high_proportion != 1) %>% -->
<!-- #   select(-is_zero_or_missing, -is_normal, -rolling_zeros_missing, -->
<!-- #          -rolling_normals, -proportion_zeros_missing, -flag_high_proportion) -->
<!-- ``` -->

<!-- ## Plot sensors with >20% zeros -->

<!-- ```{r, perc-zeros} -->
<!-- # Percentage of zero readings for each sensor -->
<!-- sensor_zero_readings <- pa_complete %>% -->
<!--   group_by(sensor_index) %>% -->
<!--   summarize(pct_zeros = round(100 * sum(pm2.5_alt == 0) / n(), 2), -->
<!--             pct_missing = round(100 * sum(pm2.5_alt == -1) / n(), 2)) %>% -->
<!--   filter(pct_zeros > 20) -->

<!-- # Loop through each sensor and create plots -->
<!-- for (i in 1:nrow(sensor_zero_readings)) { -->
<!--   s <- sensor_zero_readings$sensor_index[[i]] -->
<!--   data_sensor <- pa_complete %>% filter(sensor_index == s) -->
<!--   n <- nrow(data_sensor) -->
<!--   pz <- round(sensor_zero_readings$pct_zeros[i], 0) -->
<!--   pm <- round(sensor_zero_readings$pct_missing[i], 0) -->

<!--   # Create a data frame with segment information -->
<!--   shifted_readings <- data.frame( -->
<!--     time_stamp = head(data_sensor$time_stamp, -1), -->
<!--     time_stamp_shifted = tail(data_sensor$time_stamp, -1), -->
<!--     pm2.5_alt = head(data_sensor$pm2.5_alt, -1), -->
<!--     pm2.5_alt_shifted = tail(data_sensor$pm2.5_alt, -1) -->
<!--   ) -->

<!--   # Assign colors based on whether the pm2.5_alt value is zero or not -->
<!--   shifted_readings <- shifted_readings %>% -->
<!--     mutate(flag = ifelse(pm2.5_alt == 0, "Zero", ifelse(pm2.5_alt == -1, "Missing", "Normal"))) -->

<!--   # Plot the segments -->
<!--   p <- ggplot(data=shifted_readings, aes(x=time_stamp, xend = time_stamp_shifted, -->
<!--                                          y=pm2.5_alt, yend = pm2.5_alt_shifted, -->
<!--                                          color=flag)) + -->
<!--     geom_segment() + -->
<!--     scale_color_manual(values = c("Normal" = "black", "Missing" = "yellow", -->
<!--                                   "Zero" = "red"), -->
<!--                        name = "") + -->
<!--     labs(x = "Time", y = "PM2.5", -->
<!--          title = paste0("Sensor ", s, "\n", -->
<!--                         pz, "% zeros", "\n", -->
<!--                         pm, "% missing", -->
<!--                         "\nNumber of readings: ", n)) + -->
<!--     theme_minimal() -->

<!--   ggsave(filename = file.path("../docs", "plots", paste0( -->
<!--     "pct", pz, "_sensor", s, ".png")), -->
<!--     plot = p, width = 8, height = 6) -->
<!-- } -->
<!-- ``` -->

<!-- ## View plots of sensors with >20% zeros -->

<!-- ```{r, zero-plots, echo=FALSE, out.width="49%", out.height="20%", fig.show='hold', fig.align='center'} -->
<!-- # compare plots of different thresholds -->
<!-- img_paths <- c() -->
<!-- for (i in 1:nrow(sensor_zero_readings)) { -->
<!--   s <- sensor_zero_readings$sensor_index[[i]] -->
<!--   pz <- round(sensor_zero_readings$pct_zeros[i], 0) -->
<!--   img_path <- file.path("../docs", "plots", -->
<!--                         paste0("pct", pz, "_sensor", s, ".png")) -->
<!--   img_paths <- c(img_paths, img_path) -->
<!-- } -->
<!-- knitr::include_graphics(img_paths) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- # check which columns vary for a single sensor -->

<!-- # Group by sensor and calculate standard deviation for each column to check variability -->
<!-- sensor_variability <- purpleair_data %>% -->
<!--   group_by(sensor_index) %>% -->
<!--   summarize( -->
<!--     location_type_var = sd(location_type, na.rm = TRUE), -->
<!--     analog_input_var = sd(analog_input, na.rm = TRUE), -->
<!--     pressure_var = sd(pressure, na.rm = TRUE), -->
<!--     temperature_var = sd(temperature, na.rm = TRUE), -->
<!--     humidity_var = sd(humidity, na.rm = TRUE), -->
<!--     memory_var = sd(memory, na.rm = TRUE), -->
<!--     uptime_var = sd(uptime, na.rm = TRUE), -->
<!--     rssi_var = sd(rssi, na.rm = TRUE) -->
<!--   ) -->

<!-- # Assuming `sensor_variability` is your data.table or dataframe with variance info -->
<!-- # Calculate the mean and median of each variable's standard deviation across all sensors -->

<!-- summary_variability <- sensor_variability %>% -->
<!--   summarise(across(ends_with("_var"), list(mean = mean, median = median), na.rm = TRUE)) -->

<!-- print(summary_variability) -->
<!-- ``` -->

<!-- ## Save Processed Data to CSV -->

<!-- ```{r} -->
<!-- write.csv(purpleair_processed, file = file.path("data", "processed", "purpleair_processed.csv"), row.names = FALSE) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- # Define a helper function to get the mode (most frequent value) -->
<!-- Mode <- function(x) { -->
<!--   ux <- unique(x) -->
<!--   ux[which.max(tabulate(match(x, ux)))] -->
<!-- } -->
<!-- purpleair_processed <- purpleair_complete %>% -->
<!--   mutate( -->
<!--     time_stamp = lubridate::as_datetime(time_stamp), -->
<!--     local_timestamp = with_tz(time_stamp, tzone = "America/Los_Angeles"), -->
<!--     local_date = as.Date(local_timestamp, tz="America/Los_Angeles"), -->
<!--     dow = lubridate::wday(local_timestamp), -->
<!--     hour = lubridate::hour(local_timestamp), -->
<!--     day = lubridate::day(local_timestamp), -->
<!--     month = lubridate::month(local_timestamp), -->
<!--     year = lubridate::year(local_timestamp), -->
<!--     holiday = ifelse(local_date %in% holidays, 1, 0), -->
<!--     location_type = ifelse(is.na(location_type), # Fill `location_type` with mode for each sensor -->
<!--                            Mode(na.omit(location_type)), location_type)) %>% -->
<!--   select(-pm2.5_alt_a, -pm2.5_alt_b, -local_timestamp, -local_date) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- # Calculate and view NA percentages for each column -->
<!-- na_percentage <- purpleair_processed[, lapply(.SD, function(x) mean(is.na(x)) * 100)] -->
<!-- print(na_percentage) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- library(zoo) -->

<!-- purpleair_processed <- purpleair_processed %>% -->
<!--   group_by(sensor_index) %>% -->
<!--   arrange(time_stamp) %>% -->
<!--   mutate( -->
<!--     # Initialize imputed_flag for PM2.5 -->
<!--     imputed_flag = ifelse(is.na(pm2.5_alt), 1, 0),   -->

<!--     # Impute pm2.5_alt: Use LOCF first, then mean if any remaining NAs -->
<!--     pm2.5_alt = na.locf(pm2.5_alt, na.rm = FALSE), -->
<!--     imputed_flag = ifelse(is.na(pm2.5_alt), 1, imputed_flag),   -->
<!--     pm2.5_alt = ifelse(is.na(pm2.5_alt), mean(pm2.5_alt, na.rm = TRUE), pm2.5_alt), -->

<!--     # Impute other variables using LOCF first, then mean -->
<!--     pressure = na.locf(pressure, na.rm = FALSE), -->
<!--     pressure = ifelse(is.na(pressure), mean(pressure, na.rm = TRUE), pressure), -->

<!--     temperature = na.locf(temperature, na.rm = FALSE), -->
<!--     temperature = ifelse(is.na(temperature), mean(temperature, na.rm = TRUE), temperature), -->

<!--     humidity = na.locf(humidity, na.rm = FALSE), -->
<!--     humidity = ifelse(is.na(humidity), mean(humidity, na.rm = TRUE), humidity), -->

<!--     memory = na.locf(memory, na.rm = FALSE), -->
<!--     memory = ifelse(is.na(memory), mean(memory, na.rm = TRUE), memory), -->

<!--     uptime = na.locf(uptime, na.rm = FALSE), -->
<!--     uptime = ifelse(is.na(uptime), mean(uptime, na.rm = TRUE), uptime), -->

<!--     rssi = na.locf(rssi, na.rm = FALSE), -->
<!--     rssi = ifelse(is.na(rssi), mean(rssi, na.rm = TRUE), rssi) -->
<!--   ) %>% -->
<!--   ungroup() -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # Calculate and view NA percentages for each column -->
<!-- na_percentage <- purpleair_processed[, lapply(.SD, function(x) mean(is.na(x)) * 100)] -->
<!-- print(na_percentage) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- # Filling missing PM2.5 values with LOCF and mean imputation. -->
<!-- # Fill by first carrying forward the last observed values for each sensor.  -->
<!-- # Any remaining missing values will then be filled with the mean, ensuring no gaps remain. -->
<!-- library(zoo) # Ensure the zoo package is loaded for na.locf -->

<!-- # purpleair_processed <- purpleair_complete %>% -->
<!-- #   group_by(sensor_index) %>% -->
<!-- #   arrange(time_stamp) %>% -->
<!-- #   mutate( -->
<!-- #     pm2.5_alt = na.locf(pm2.5_alt, na.rm = FALSE),       # Use LOCF for PM2.5 values -->
<!-- #     pm2.5_alt = ifelse(is.na(pm2.5_alt), mean(pm2.5_alt, na.rm = TRUE), pm2.5_alt)  # Fill remaining NAs with mean -->
<!-- #   ) %>% -->
<!-- #   ungroup() -->

<!-- purpleair_processed <- purpleair_processed %>% -->
<!--   group_by(sensor_index) %>% -->
<!--   arrange(time_stamp) %>% -->
<!--   mutate( -->
<!--     imputed_flag = ifelse(is.na(pm2.5_alt), 1, 0),  # Initialize flag for missing values -->
<!--     pm2.5_alt = na.locf(pm2.5_alt, na.rm = FALSE),  # Use LOCF to fill missing values -->
<!--     imputed_flag = ifelse(is.na(pm2.5_alt), 1, imputed_flag), # Update flag for remaining NAs -->
<!--     pm2.5_alt = ifelse(is.na(pm2.5_alt), mean(pm2.5_alt, na.rm = TRUE), pm2.5_alt) # Fill remaining NAs with mean -->
<!--   ) %>% -->
<!--   ungroup() -->
<!-- ``` -->


