---
title: "Download OSM data"
output: 
  github_document:
    output_dir: "docs"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.path = "docs/plots/")

osm_directory <- readr::read_file("inputs/osm_directory.txt")
purpleair_directory <- readr::read_file("inputs/purpleair_directory.txt")
preprocessing_directory <- readr::read_file("inputs/preprocessing_directory.txt")
sensor_directory <- paste0(osm_directory,"/sensor_data")
```

## Load required libraries
```{r, load-libraries, message = FALSE, warning = FALSE}
library(osmdata) # Open Street Map
library(dplyr) # For data manipulation
library(data.table) # Faster than dataframes (for big files)
library(sf) # For working with spatial data
library(leaflet) # For interactive maps
library(tigris) # Counties map data
library(ggplot2)
```

## Read files
```{r, read-files, eval=TRUE}
purpleair_data <- fread(paste0(preprocessing_directory,"/purpleair_filtered_2018-2019.csv"))
purpleair_sensors <- st_read(paste0(preprocessing_directory, "/pasensors_weatherstations.gpkg"), quiet = TRUE)
```

## Filter PurpleAir sensors
```{r, filter_pa_sensors, eval=TRUE}
# filter purpleair_sensors for our selected data
filtered_sensors <- purpleair_sensors %>% 
  filter(sensor_index %in% unique(purpleair_data$sensor_index))
```

## Create Buffers around Purple Air Sensors
```{r, pa-buffers, eval=TRUE}
# change crs to be able to create buffers
crs_ca <- 3310
filtered_sensors <- st_transform(filtered_sensors, crs_ca)

# buffer radius in meters
buffer <- 1000
purpleairs_buffers <- st_buffer(filtered_sensors, dist=buffer, nQuadSegs=2)

# change crs back to original
crs <- 4326
filtered_sensors <- st_transform(filtered_sensors, crs)
purpleairs_buffers <- st_transform(purpleairs_buffers, crs)
```

# Map of PurpleAir sensors with buffers
```{r, pa-map, eval=TRUE, message=FALSE, results = FALSE}
ca <- counties("California", cb = TRUE, progress_bar=FALSE)

ggplot() +
  geom_sf(data = ca, color="black", fill="antiquewhite", size=0.25) +
  geom_sf(data = purpleairs_buffers, fill = "lavender") +
  geom_sf(data = filtered_sensors, color = "purple", size = 0.1) +
  coord_sf(xlim = c(-122.55, -122.35), ylim = c(37.83, 37.7)) +
  theme(panel.background = element_rect(fill = "aliceblue")) +
  xlab("Longitude") + ylab("Latitude") +
  ggtitle("PurpleAir in San Francisco")
```
```{r}
filtered_sensors
purpleairs_buffers <- purpleairs_buffers %>% filter(sensor_index %in% z$sensor_index)
```


# Download OSM roads surrounding each sensor
```{r, download-roads, eval=TRUE, warning=FALSE}
# Loop through each sensor buffer and download OSM data
for (i in 1:nrow(purpleairs_buffers)) {
  output_name <- paste0('sensor',purpleairs_buffers[i,]$sensor_index)
  filename <- paste0(sensor_directory,"/", output_name, "_roads_osm.gpkg")
  
  # if file exists, skip it
  if (file.exists(filename)) {
    next
  }
  
  osm <- opq(bbox = purpleairs_buffers[i,]$geom) %>%
    add_osm_feature(key = 'highway') %>%
    osmdata_sf()
  
  # if cell is on empty location skip it
  if (is.null(osm$osm_lines)) {
    next
  }
  
  # Select only the columns you want to keep (if any col doesnt exist fill with NA)
  # if column osm_id is missing fill it with rownames
  if(!"osm_id" %in% names(osm$osm_lines)) {
    osm$osm_lines$osm_id <- rownames(osm$osm_lines)
  }
  # if column name is missing fill it with NA
  if(!"name" %in% names(osm$osm_lines)) {
    osm$osm_lines$name <- NA
  }  
  # if column highway is missing fill it with NA
  if(!"highway" %in% names(osm$osm_lines)) {
    osm$osm_lines$highway <- NA
  }  
  # if column lanes is missing fill it with NA
  if(!"lanes" %in% names(osm$osm_lines)) {
    osm$osm_lines$lanes <- NA
  }
  # if column maxspeed is missing fill it with NA
  if(!"maxspeed" %in% names(osm$osm_lines)) {
    osm$osm_lines$maxspeed <- NA
  }
  
  selected_columns <- osm$osm_lines %>% select(osm_id, name, highway, lanes, maxspeed)

  # Create an sf object
  sf_obj <- st_as_sf(selected_columns)
  
  # get intersection with buffer
  sf_obj <- st_intersection(sf_obj, purpleairs_buffers[i,]$geom)
  
  # if intersection is empty
  if (is.null(sf_obj) || nrow(sf_obj) == 0) {
    next
  }
  # Add column for sensor index
  sf_obj$sensor_index <- purpleairs_buffers[i,]$sensor_index
  
  # Save the sf object as a shapefile
  st_write(sf_obj, filename, driver = "GPKG", append=FALSE, quiet=TRUE)
}
```

## Read sensor roads files and save to one file
```{r, save-road-to-one-file, eval=TRUE, message=FALSE, warning=FALSE}
# Get a list of file paths
file_paths <- list.files(sensor_directory, pattern = "^sensor.*_roads_osm\\.gpkg$", full.names = TRUE)

# Read all shapefiles into a list
sf_list <- lapply(file_paths, function(file_path) {
  sf::st_read(file_path, quiet = TRUE)
})

# Merge the spatial objects
merged_sf <- do.call(rbind, sf_list)

# Drop duplicate rows
merged_sf <- merged_sf %>% distinct()

# Write the merged spatial object to a new shapefile
st_write(merged_sf, paste0(osm_directory,"/","bayarea_roads_osm.gpkg"), driver = "GPKG", append=FALSE, quiet=TRUE)
```

# Download OSM buildings surrounding each sensor
```{r, download-buildings, eval=TRUE, warning=FALSE}
# Loop through each sensor buffer and download OSM data
for (i in 1:nrow(purpleairs_buffers)) {
  output_name <- paste0('sensor',purpleairs_buffers[i,]$sensor_index)
  filename <- paste0(sensor_directory,"/", output_name, "_buildings_osm.gpkg")
  # if file exists, skip it
  if (file.exists(filename)) {
    next
  }
  osm <- opq(bbox = purpleairs_buffers[i,]$geom) %>%
    add_osm_feature(key = 'building') %>%
    osmdata_sf()
  
  # if cell was on an empty location
  if (is.null(osm$osm_polygons) || nrow(osm$osm_polygons) == 0) {
    next
  }
  
  # Select only the columns you want to keep (if any col doesnt exist fill with NA)
  
  # if column osm_id is missing fill it with rownames
  if(!"osm_id" %in% names(osm$osm_polygons)) {
    osm$osm_polygons$osm_id <- rownames(osm$osm_polygons)
  }
  # if column name is missing fill it with NA
  if(!"name" %in% names(osm$osm_polygons)) {
    osm$osm_polygons$name <- NA
  }  
  # if column building is missing fill it with NA
  if(!"building" %in% names(osm$osm_polygons)) {
    osm$osm_polygons$building <- NA
  }  
  # if column amenity is missing fill it with NA
  if(!"amenity" %in% names(osm$osm_polygons)) {
    osm$osm_polygons$amenity <- NA
  }
  
  selected_columns <- osm$osm_polygons %>% select(osm_id, name, building, amenity)

  # Create an sf object
  sf_obj <- st_as_sf(selected_columns)
  
  # get intersection with buffer
  sf_obj <- st_intersection(sf_obj, purpleairs_buffers[i,]$geom)
  
  # if intersection is empty
  if (is.null(sf_obj) || nrow(sf_obj) == 0) {
    next
  }
  
  # Add column for sensor index
  sf_obj$sensor_index <- purpleairs_buffers[i,]$sensor_index
  
  # Save the sf object as a shapefile
  st_write(sf_obj, filename, driver = "GPKG", append=FALSE, quiet=TRUE)
}
```

## Read sensor buildings files and save to one file
```{r, save-buildings-to-one-file, eval=TRUE, message=FALSE, warning=FALSE}
# Get a list of file paths
file_paths <- list.files(sensor_directory, pattern = "^sensor.*_buildings_osm\\.gpkg$", full.names = TRUE)

# Read all shapefiles into a list
sf_list <- lapply(file_paths, function(file_path) {
  sf::st_read(file_path, quiet = TRUE)
})

# Merge the spatial objects
merged_sf <- do.call(rbind, sf_list)

# Drop duplicate rows
merged_sf <- merged_sf %>% distinct()

# Write the merged spatial object to a new shapefile
st_write(merged_sf, paste0(osm_directory,"/","bayarea_buildings_osm.gpkg"), driver = "GPKG", append=FALSE, quiet=TRUE)
```

# Download OSM trees surrounding each sensor
```{r, download-trees, eval=TRUE, warning=FALSE}
# Loop through each sensor buffer and download OSM data
for (i in 1:nrow(purpleairs_buffers)) {
  output_name <- paste0('sensor',purpleairs_buffers[i,]$sensor_index)
  filename <- paste0(sensor_directory,"/", output_name, "_trees_osm.gpkg")
  
  # if file exists, skip it
  if (file.exists(filename)) {
    next
  }

  osm <- opq(bbox = purpleairs_buffers[i,]$geom) %>%
    add_osm_feature(key = 'natural') %>%
    osmdata_sf()
  
  tree_points <- osm$osm_points[!is.na(osm$osm_points$natural), ]
  tree_points <- tree_points[tree_points$natural == "tree", ]
  
  # if cell was on an empty location
  if (is.null(tree_points) || nrow(tree_points) == 0) {
    next
  }
  
  # Select only the columns you want to keep (if any col doesnt exist fill with NA)
  
  # if column osm_id is missing fill it with rownames
  if(!"osm_id" %in% names(tree_points)) {
    tree_points$osm_id <- rownames(tree_points)
  }
  
  selected_columns <- tree_points %>% select(osm_id)

  # Create an sf object
  sf_obj <- st_as_sf(selected_columns)
  
  # get intersection with buffer
  sf_obj <- st_intersection(sf_obj, purpleairs_buffers[i,]$geom)
  
  # if intersection is empty
  if (is.null(sf_obj) || nrow(sf_obj) == 0) {
    next
  }
  # Add column for sensor index
  sf_obj$sensor_index <- purpleairs_buffers[i,]$sensor_index
  
  # Save the sf object as a shapefile
  st_write(sf_obj, filename, driver = "GPKG", append=FALSE, quiet=TRUE)
}
```

## Read sensor trees files and save to one file
```{r, save-trees-to-one-file, eval=TRUE, message=FALSE, warning=FALSE}
# Get a list of file paths
file_paths <- list.files(sensor_directory, pattern = "^sensor.*_trees_osm\\.gpkg$", full.names = TRUE)

# Read all shapefiles into a list
sf_list <- lapply(file_paths, function(file_path) {
  sf::st_read(file_path, quiet = TRUE)
})
# Merge the spatial objects
merged_sf <- do.call(rbind, sf_list)

# Drop duplicate rows
merged_sf <- merged_sf %>% distinct()

# Write the merged spatial object to a new shapefile
st_write(merged_sf, paste0(osm_directory, "/", "bayarea_trees_osm.gpkg"), driver = "GPKG", append=FALSE, quiet=TRUE)
```

## Read merged osm files
```{r, read-buildings-roads-trees, message=FALSE, warning=FALSE}
# read roads file
bayarea_roads <- st_read(paste0(osm_directory, "/bayarea_roads_osm.gpkg"), quiet = TRUE)

# read buildings file
bayarea_buildings <- st_read(paste0(osm_directory, "/bayarea_buildings_osm.gpkg"), quiet = TRUE)

# read trees file
bayarea_trees <- st_read(paste0(osm_directory, "/bayarea_trees_osm.gpkg"), quiet = TRUE)

# Load california counties for mapping
ca <- counties("California", cb = TRUE, progress_bar=FALSE)
```

## Plot roads
```{r, plot-roads, results = FALSE}
ggplot() + 
  geom_sf(data = ca, color="black", fill="antiquewhite", size=0.25) +
  geom_sf(data = bayarea_roads, color = "cornflowerblue", size = 0.1) +
  coord_sf(xlim = c(-123.8, -121.0), ylim = c(36.9, 39.0)) +
  theme(panel.background = element_rect(fill = "aliceblue")) + 
  xlab("Longitude") + ylab("Latitude") +
  ggtitle("OSM Roads")
```

## Plot buildings
```{r, plot-buildings, results = FALSE}
ggplot() + 
  geom_sf(data = ca, color="black", fill="antiquewhite", size=0.25) +
  geom_sf(data = bayarea_buildings, color="darkorange3", fill = "darkorange") +
  coord_sf(xlim = c(-123.8, -121.0), ylim = c(36.9, 39.0)) +
  theme(panel.background = element_rect(fill = "aliceblue")) + 
  xlab("Longitude") + ylab("Latitude") +
  ggtitle("OSM Buildings")
```

## Plot trees
```{r, plot-trees, results = FALSE}
ggplot() + 
  geom_sf(data = ca, color="black", fill="antiquewhite", size=0.25) +
  geom_sf(data = bayarea_trees, color = "darkgreen", fill="forestgreen", size = 0.1) +
  coord_sf(xlim = c(-123.8, -121.0), ylim = c(36.9, 39.0)) +
  theme(panel.background = element_rect(fill = "aliceblue")) + 
  xlab("Longitude") + ylab("Latitude") +
  ggtitle("OSM Trees")
```

## Select small area of San Francisco to map
```{r, map-smaller-area, message=FALSE, warning=FALSE}
crs = 4326
bbox <- c(xmin = -122.46, ymin = 37.76, xmax = -122.42, ymax = 37.72)
bbox_polygon <- st_as_sfc(st_bbox(bbox))
st_crs(bbox_polygon) <- crs
sanfran_roads <- st_intersection(bayarea_roads, bbox_polygon)
sanfran_buildings <- st_intersection(bayarea_buildings, bbox_polygon)
sanfran_trees <- st_intersection(bayarea_trees, bbox_polygon)
sanfran_sensors <- st_intersection(filtered_sensors, bbox_polygon)
sanfran_buffers <- st_intersection(purpleairs_buffers, bbox_polygon)
```

## Plot san fran city
```{r, map-city, warning=FALSE}
leaflet() %>%
  # addPolylines(data = bbox_polygon, color = "black", weight = 3, opacity = 1, popup = "Bounding Box") %>%
  addPolylines(data = sanfran_roads, color = "cornflowerblue", weight = 1, popup = "Roads") %>%
  addPolygons(data = sanfran_buildings, fillOpacity = 1, weight = 1, color="darkorange", fillColor = "orange", popup = "Buildings") %>%
  addCircleMarkers(data = sanfran_trees, popup = "Trees",
                   fillColor = "#99CC99", fillOpacity = 1,
                   color = "#336600", weight=1, opacity = 1, radius = 1.5) %>%
  addCircleMarkers(data = sanfran_sensors, popup = "Sensors",
                   fillColor = "#CC66CC", fillOpacity = 1,
                   color = "#9933CC",weight=2, opacity = 1, radius = 5) %>%
  addPolygons(data = sanfran_buffers, color = "#9933CC", weight = 1, opacity = 1, fillOpacity = 0, popup = "Buffers") %>%
  addProviderTiles("CartoDB") %>% 
  addLegend(colors = c("#9933CC", "cornflowerblue", "darkorange", "#336600"), 
            labels = c("PurpleAir Sensors", "Roads", "Buildings", "Trees"))
```

## Plot san fran city
```{r, map-city, warning=FALSE}

library(htmlwidgets)
library(htmltools)


major_roads <- c("motorway", "motorway_link", "trunk", "primary")

# group road types to major and minor
osm_roads_pa <- bayarea_roads %>%
  mutate(highway = ifelse(is.na(highway), "NA", highway)) %>%
  mutate(road_type = ifelse(highway %in% major_roads, "Major", "Minor"),
         road_length = round(st_length(geom),2)) 

# Define building categories
house <- c("house", "detached", "semidetached_house", "houses", "farm")
apartments <- c("residential", "apartments")
undefined <- c("NA", "yes")

# Categorize buildings
buildings <- bayarea_buildings %>% 
  mutate(building = ifelse(is.na(building), "NA", building)) %>%
  mutate(building_type = case_when(
    building %in% undefined ~ "Undefined",
    building %in% house ~ "House",
    building %in% apartments ~ "Apartments",
    TRUE ~ "Other"
  ),
  building_area = round(st_area(geom),2))

# Remove units from road_length
# attributes(road_lengths$road_length) = NULL

# Define colors for road types
road_colors <- colorFactor(c("cornflowerblue", "lightblue"), domain = c("Major", "Minor"))
building_colors <- colorFactor(c("orangered", "darkorange", "peachpuff", "papayawhip"), 
                               domain = c("Other", "Apartments", "House", "Undefined"))

m <- leaflet() %>%
  addPolylines(data = osm_roads_pa, color = ~road_colors(road_type), weight = 2,
               label = ~HTML(paste(htmlEscape(road_type), " (", htmlEscape(highway),
                              ")<br>Length: ", htmlEscape(road_length), " [m]")),
               popup = "Roads") %>%
  addPolygons(data = buildings, fillOpacity = 1, weight = 1,
              color=~building_colors(building_type), fillColor = ~building_colors(building_type),
              label = ~HTML(paste(htmlEscape(building_type), " (", htmlEscape(building),
                             ")<br>Area: ", htmlEscape(building_area), " [m^2]"),
              popup = "Buildings")) %>%
  addCircleMarkers(data = bayarea_trees, popup = "Trees", label = "Tree",
                   fillColor = "#99CC99", fillOpacity = 1,
                   color = "#336600", weight=1, opacity = 1, radius = 2) %>%
  addCircleMarkers(data = z, popup = "Sensors",
                   label = ~paste("Sensor:", htmlEscape(sensor_index)),
                   fillColor = "#CC66CC", fillOpacity = 1, color = "#9933CC",
                   weight=2, opacity = 1, radius = 5) %>%
  addPolylines(data = purpleairs_buffers, color = "#9933CC", weight = 1,
              opacity = 1, popup = "Buffers") %>%
  addProviderTiles("CartoDB") %>% 
  addLegend(colors = c("#9933CC", "#336600", "cornflowerblue", "lightblue",
                       "orangered", "darkorange", "peachpuff", "oldlace"), 
            labels = c("PurpleAir Sensors", "Trees", "Major Roads", "Minor Roads",
                       "Other", "Apartments", "House", "Undefined"))
m
saveWidget(m, file="san_fran_map.html")
```
